### 2021-11-23 minta.tex
Initial Overleaf Import

\documentclass{llncs}

\usepackage[a4paper,includeheadfoot,top=1.65in,bottom=1.65in,left=1.73in,hcentering]{geometry}
\usepackage[pdftex]{graphicx}
\usepackage[sectionbib]{natbib}

\usepackage[utf8]{inputenc}
\usepackage{t1enc}
\usepackage{tikz-dependency}
\usepackage{qtree}
\usepackage{float}
\restylefloat{figure}
\usepackage{tikz-qtree}
\usepackage[english]{babel}
\usepackage{todonotes}
\selectlanguage{english}

\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{soul}

\newcommand{\embert}{\texttt{emBERT}}
\newcommand{\emtsv}{\texttt{emtsv}}
\newcommand{\hubert}{\texttt{huBERT}}
\newcommand{\magyarlanc}{\texttt{magyarlanc}}
\newcommand{\spacy}{spaCy}
\newcommand{\udpipe}{UDPipe}
\newcommand{\stanza}{Stanza}
\newcommand{\huspacy}{HuSpaCy}
\newcommand{\huntoken}{HunToken}
\newcommand{\lemmy}{Lemmy}

\renewcommand\todo{\noop}
% \renewcommand\bibname{References}

\begin{document}

%\pagestyle{myheadings}
%\def\leftmark{{\rm XIV. Magyar Sz\'am\'\i t\'og\'epes Nyelv\'eszeti Konferencia}}
%\def\rightmark{{\rm Szeged, 2018. január 18-19.}}

%  \setcounter{page}{1}

\title{\huspacy: an industrial-strength Hungarian natural language processing toolkit}
%a november 15-ig bekuldott verziobol kerjuk a szerzok adatait kihagyni
%	\author{Orosz György\inst{1}, Farkas Richárd\inst{1}, Vincze Veronika\inst{1,2}\\
% \institute{$^1$Szegedi Tudom\'anyegyetem, Informatikai Int\'ezet\break
%Szeged, \'Arp\'ad t\'er 2.\break
%$^2$MTA-SZTE Mesters\'eges Intelligencia Kutat\'ocsoport\break
%Szeged, Tisza Lajos k\"or\'ut 103.\break 
%\{zsibrita,rfarkas,vinczev\}@inf.u-szeged.hu\break
%}
%}

\maketitle

\begin{abstract}
Although there are a couple of open-source language processing pipelines available for Hungarian, none of them satisfies the requirements of today’s NLP applications. A language processing pipeline should consist of close to state-of-the-art lemmatization, morphosyntactic analysis, entity recognition and word embeddings. 
Industrial text processing applications have to satisfy non-functional software quality requirements, what is more, frameworks supporting multiple languages are more and more favored.
%have been released in large numbers in recent years, such pipelines also have to satisfy non-functional software quality requirements.  
%Last, but not least, the international research and industrial community have been experimenting with languages whose language-specific analyzer tools are available in multilingual language processing libraries. 
This paper introduces \huspacy, an industry-ready Hungarian language processing pipeline. The presented tool provides components for the most important basic linguistic analysis tasks. It is open-source and is available under a permissive license. Our system is built upon spaCy’s NLP components which means that it is fast, has a rich ecosystem of NLP applications and extensions, comes with extensive documentation and a well-known API. Besides the overview of the underlying models, we also present rigorous evaluation on common benchmark datasets. Our experiments confirm that HuSpaCy has high accuracy in all subtasks while maintaining resource-efficient prediction capabilities. 
\end{abstract}

\section{Introduction}

Basic natural language processing tasks such as tokenization, sentence splitting, part-of-speech tagging, lemmatization, dependency parsing, named entity recognition and word embedding representations are amongst the most widely studied problems in natural language processing. Several text analysis applications have been developed during the last decades for both English and other less-resourced languages such as Hungarian. However, a large majority of them solely focus on achieving high accuracy scores on artificial benchmarks and ignore the importance of the underlying model’s practical usability.

In this paper we introduce \huspacy, an industry-strength Hungarian text processing pipeline capable of parsing and tagging texts with high accuracy on limited computational resources. Our system is built upon \spacy’s\footnote{\url{https://spacy.io/}} NLP components, which means that it is fast, has a rich ecosystem of NLP applications and extensions, comes with extensive documentation and a well-known API.

Besides the overview of the underlying models, we also  present rigorous evaluation using the Universal grammatical annotations. Our experiments confirm that \huspacy has near state-of-the-art accuracy in many subtasks while maintaining resource-efficient prediction capabilities.

\section{Background}

\subsection{Demands for a language processing pipeline in the 2020s}
Starting from the release of the Penn Treebank \citep{penntreebank} in 1992, the research community developed language processing tools for particular tasks, like tokenization, part-of-speech tagging etc. These tools are usually run in a sequence and form a pipeline. In the 2000s, many language-specific corpora and treebanks were developed along with such pipelines. Hungarian was among the best supported languages \citep{metanet} ten years ago.

In the early 2010s, Universal PoS \citep{udpos} and Universal Dependency \citep{univedepv1} labeling schemata were developed with the goals of ''cross-linguistically consistent treebank annotation for many languages'' and ''facilitating multilingual parser development, cross-lingual learning, and parsing research from a language typology perspective.'' Many language-specific pipelines changed their representations to the universal annotation schema, but most of them stayed in their own software architecture. Industrial NLP applications are frequently multi-lingual, i.e. the same NLP task has to be solved in several languages. The demand for standardization over languages is high in commercial partners. Beyond universal PoS and dependency annotations, companies who are not NLP experts but want to apply language processing tools prefer multilingual software frameworks and make business decisions to support Hungarian, based on the availability in multilingual frameworks.

The last five years of NLP are dominated by neural language models (NLM) and the applications based on them \citep{nlp-trends}. Academic research has introduced various deep learning methods outperforming previous state of the art in many areas. Such systems usually employ a single neural network providing end-to-end NLP solutions without the need for specific pipeline steps. Also, several pre-trained multilingual NLMs are becoming available which provide standardised solutions for many languages. Regardless of the multilinguality and the high accuracy of deep learning solutions, a lot of critiques have been raised by real-world industrial NLP projects recently.

These are as follows: deep learning solutions often require far more computational resources compared to classic solutions. They heavily rely on GPU acceleration along with significant memory consumption. What is more, their running costs are usually 10 or even 100 times higher than that of alternative solutions. These drawbacks are questioning the commercial return of the accuracy gain. We also note that modern NLP pipelines consist of static word embedding representations and use deep learning for individual pipeline steps as well, hence, the advantage of large neural end-to-end systems
%\footnote{e.g. BERT models} 
are very small.

Another industrial demand about language processing systems is to provide human-readable output. Most of the industrial applications are fully or partially rule-based solutions, as (enough) training data for a pure machine learning solution is not available. And there is no free lunch! Each and every real-world application has its own requirements. Rule-based components of these real-world applications require language-specific representations which can be used for defining rules. Such human-readable representations consist of tokens, lemmata, part-of-speech tags, morphological features, dependency parse trees and named entities. Static word embeddings are often integral parts of industrial applications, as many practical algorithms (e.g. semantic textual similarity methods) heavily rely on them.

\subsection{Requirements for industrial-strength language processing pipelines}

Considering decades of experience of practical NLP applications, developing an “industrial strength” text processing system is a challenging task. First of all, such a tool should tackle the most important text preprocessing tasks including tokenization, sentence splitting, PoS tagging, lemmatization, dependency parsing, named entity recognition and word embedding representation.

Next, the application has to be accurate enough for real world scenarios while it should be resource conscious at the same time. Furthermore, an industry focused system should be developer friendly, customizable and easy-to-integrate, as NLP modules are integral parts of a larger system in practical applications. These requirements imply that solid documentation should be available as well. Moreover, it is often desired that the underlying machine learning model(s) should be reproducible and controllable.

Last but not least, modern NLP applications are usually multilingual, thus compatibility with international annotation standards  \citep{udpos,univedepv1,UniDep,unimorph} is necessary. Moreover, it is also preferred to be easily usable through a well-known multilingual toolkit.

\subsection{The landscape of Hungarian language processing pipelines}

Up until recently, only a few text processing applications were focused on meeting these criteria even for English. When spaCy was released in 2015 \citep{spacy}, it was one one of the first tools targeting mainly real-world industrial applications. Its authors created an unprecedented tool which offers near state-of-the-art accuracy while being an order of magnitude faster than other tools available. SpaCy also comes with an intuitive API, has detailed documentation, and also fits well into the Python ecosystem of machine learning tools. What is more, it is easily deployable and offers built-in syntax and entity visualization tools as well.

The landscape of the Hungarian text processing systems is similar to that of English before the ''industrial NLP revolution''. There are a number of standalone text analysis tools\footnote{cf. \url{https://github.com/oroszgy/awesome-hungarian-nlp}} \citep{metanet} capable of performing individual text processing tasks, but they often do not play well with each other.

In contrast, there are only a few attempts at providing industrial Hungarian pipelines. One of them is \magyarlanc{} \citep{magyaralanc} which is a Java based system consisting of state-of-the-art pipeline steps, which were adapted and extended from various libraries. It was designed to serve industrial applications, what is more, a lot of effort has been made on software quality, speed, memory efficiency and customizability. It performs tokenization, sentence boundary detection (SBD), PoS tagging, lemmatization and dependency parsing, but lacks entity recognition and word embeddings. It uses the version 1 of the Universal Dependency (UD) annotations. Although the tool is still used in real world commercial applications, it is not maintained for years.

There is only one other attempt to provide a unified framework for Hungarian text processing tools: \emtsv{} \citep{emtsv1, emtsv2, emtsv3} (and its predecessor e-magyar \citep{emagyar1, emagyar2}) is a result of a multi-institute collaboration project aiming to integrate existing NLP toolkits into a single application. Unfortunately, neither computational efficiency nor developer ergonomics were amongst the main goals of the project. Although \emtsv{} can yield Universal morphosyntactic annotations through conversion, it is rather inaccurate. What is more, it is not designed to efficiently deal with word vectors, therefore no such facility is available in its system.

Talking of Hungarian-specific pipelines, we must mention the contenders of the recent multilingual CoNLL text parsing competitions \citep{conll-2017, conll-2018}. There were numerous  submissions, but \stanza{} \citep{stanza} and \udpipe{} \citep{udpipe} are by far the most popular freely available off-the-shelf applications. These tools provide morphological and syntactical analysis of raw texts for many languages, but lack entity annotations. Accuracy scores vary across tools, but all of them are limited by the small size of the publicly available UD annotated gold standard corpora.

\newlength{\lsys}
\settowidth{\lsys}{embeddings}
\begin{table}
\begin{center}
\begin{tabular}{
%lccccc
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
  >{\centering\arraybackslash}m{\lsys}
}
\toprule
           & NER & 
           \begin{tabular}{@{}c@{}}Word  \\ embeddings\end{tabular} &
           \begin{tabular}{@{}c@{}}High \\ throughput\end{tabular} & 
           \begin{tabular}{@{}c@{}}Part of a \\ multilingual \\ pipeline\end{tabular}
             &
             \begin{tabular}{@{}c@{}}Free for \\ commercial \\ usage\end{tabular} \\
\midrule
\magyarlanc{} & -                       & -                                   & \checkmark                                   & -                                                   & \checkmark                                             \\
\emtsv{}      & \checkmark                       & -                                   & -                                   & -                                                   & -                                             \\
\udpipe{}     & -                       & -                                   & \checkmark                                   & \checkmark                                                   & -                                             \\
\stanza{}     & \checkmark                       & \checkmark                                   & -                                   & \checkmark                                                   & \checkmark                                            
\\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Hungarian language processing pipelines evaluated with regards of requirements of industrial applicability}
\label{table:systems}
\end{center}
\end{table}

Table \ref{table:systems} summarizes the landscape of the most important Hungarian language processing pipelines and show how they meet the requirements of today’s NLP applications. All in all, we can say that none of them is easily applicable in industrial settings.

We present \huspacy{}, a new industry-ready Hungarian natural language processing toolkit\footnote{We implemented our pipeline in spaCy v3.x. Prior to this work, there were attempts (\url{https://github.com/oroszgy/spacy-hungarian-models}) to support Hungarian in this framework, but none of those releases are compatible with the tool’s current version.}. 
It provides all the aforementioned basic text processing modules with high accuracy. The underlying models are optimized to be light on memory consumption and CPU usage. The presented tool is open source\footnote{The URL is to be shared in the final paper.} and is available under the permissive CC-BY-SA-4.0 license. Our system is built on top of spaCy’s infrastructure, thus extensive documentation, debugging tools, an ergonomic API and a flourishing ecosystem are already provided.

\section{Pipeline steps}

This section introduces the NLP algorithms behind \huspacy{}. The presented system is built on spaCy’s architecture, therefore we mainly relied on its symbolic and ML-based text processing infrastructure. The following paragraphs give only a high-level overview of the internals of the parent system and rather concentrate on the contributions of this work.

\subsection{Tokenization}

As regards tokenization, the developed solution uses and extends the underlying system’s algorithm \citep{spacy-tokenization}. It works as follows: first the input text is split on whitespaces, then token boundaries are identified by splitting prefixing or suffixing character sequences. To make this algorithm viable for Hungarian, we extended it with language specific prefix and suffix splitting rules. Furthermore, we had to deal with the ambiguity of tokens around the full stop character, thus a decent abbreviation list has been incorporated to increase the module’s accuracy. During this process we mostly relied on the tests of \huntoken{} \citep{huntoken}.

\subsection{Morphosyntactic tagging, sentence splitting and parsing}
\label{sec:tagging}

Sentence boundaries, dependency parse trees, PoS tags, and the corresponding morphosyntactic features are predicted by a multitask deep learning model of the underlying NLP framework. SpaCy’s machine learning approach can be summarized as ''embed, encode, attend, predict''  \citep{spacy-neural-model} which our system adapts for its tagging and parsing components.

Tokens are embedded using the concatenation of static (pretrained) word vectors and ones learned during the task-specific training process. We use a publicly\footnote{\url{https://github.com/oroszgy/hunlp-resources/releases/tag/webcorpuswiki_word2vec_v0.1}} available 300d word embedding which has been trained on the Hungarian Webcorpus \citep{HunWebCorpus} and a snapshot of the Hungarian Wikipedia with CBOW methodology \citep{word2vec}. Task specific word vectors are 256 wide consisting 64 dimensional embeddings of the tokens’ prefixes, suffixes, shapes and the lowercase forms. To make such computation efficient, feature hashing is extensively applied to all kinds of input strings.

During the encoding part, vectors are passed through a four deep stacked CNN encoder \citep{cnn} which uses residual connections and is accompanied with maxout pooling\footnote{The pooling step is considered to be the “attention” mechanism.} \citep{spacy-parser2}. The tagger predicts supertags for tokens which are the combinations of the dependency, part-of-speech and morphosyntactic labels. Efficient prediction is guaranteed by the underlying greedy tagger consisting only of a linear and a softmax layer. As for the dependency parsing, an arc-eager transition system \citep{spacy-parser} is utilized, which shares weights with the tagger model through multitask learning.

Finally, sentence boundary recognition is formalized as a sequence tagging problem where tokens are tagged with a binary label indicating the first token of a sentence. This component is an integral part of the multitask architecture, thus it also shares its lower layers with the parser and the morphosyntactic tagger.

\subsection{Lemmatization}

SpaCy’s default lemmatization model is mainly designed for English. It is not suitable for morphologically complex languages such as Hungarian as it only uses lookup tables. Hence, we decided to look for a more sophisticated solution, and integrated the \lemmy{} toolkit \citep{lemmy} which is an open-source Python implementation of the CST rule-learning engine \citep{cst-lemmatizer}. To improve its accuracy we incorporated three minor modifications. First, prefixing numbers of numeric tokens are masked to help the generalization of the engine for inflected numbers. (For example the masked token of `2021-ben’ becomes `0000-ben’.) Second, we enforce lowercase lemmata of sentence starting tokens if they are not proper nouns. Finally, if there are multiple lemma candidates available for a given (word, tag) pair, we pick the one with the highest frequency on the training dataset.

\subsection{Named entity recognition}

SpaCy’s entity recognizer is built on the transition-based parser architecture described in Section \ref{sec:tagging} (similarly to \citet{spacy-ner}). However, there are two key differences compared to the system of \citet{spacy-ner} The first is that the set of possible transition actions reflects the BILOU tagging scheme. This trick allows the model to have better discrimination ability between different entity classes, furthermore it makes the learning problem easier. Second, the state vector computation includes clues not just from the surrounding words but the tokens of previous entities as well. The sequence tagger model uses BILOU tags for encoding entity boundaries and the decoder is based on a simple greedy softmax layer similar to the morphosyntactic tagger.

\section{Text parsing experiments}

In order to show \huspacy{} abilities, we performed a series of experiments and compared it with the most popular Hungarian NLP pipelines available. For each component, we measured its performance on benchmark datasets along with the model’s throughput and memory consumption. As named entity recognition is usually not part of such pipelines, first we focus on evaluating the text parsing task including tokenization, sentence splitting, morphosyntactic tagging, lemmatization and dependency parsing.

We compare our system with the Hungarian specific \emtsv{} pipeline and two other multilingual toolkits, namely \udpipe{} and \stanza{}. The latter tools score high on parsing UD corpora and both are off-the-shelf available systems. We used these tools as black boxes meaning that no efforts have been made to retrain or fine-tune them. Evaluation is carried out on the test set of the Hungarian subset of the Universal Dependencies Corpus \citep{UniDep} which consists of a small, manually transcribed part of the Szeged Corpus. Performance measures are provided by the evaluation script of the CoNLL 2018 Shared Task\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.

Up until now, there has  only been a single Hungarian corpus \citep{szegedcorpus} having both morphosyntactic and dependency parse annotations. What is more, UD annotations are available only in a rather small subcorpus of it \citep{UniDep}. While morphosyntactic labels can be transcribed automatically from the Hungarian-specific formalism to UD with high accuracy, the same is not true for dependency parse trees. Hence, there is a notable difference in size of the training data available for tagging and parsing.

To best utilize all possible data sources, we applied a two-step learning strategy
%\footnote{Our experiments are fully reproducible. Hyperparameters of the models lie in configuration files of the given repository.}
using \huspacy\textquotesingle s multitask deep learning model. In the first step, the tagger and the SBD components are pre-trained on the full-size transcribed Szeged Corpus\footnote{When we refer to the Szeged Corpus as a training set, we mean all the sentences that are not part of the development or the test set of the Universal Dependencies corpus.}. This is followed by a fine-tuning step on the gold standard dataset where dependency annotations are also learned. A single step model relying solely on the UD data is also involved in our experiments to allow fair comparison with other tools. 
%%\hl{A summary of hyperparameters can be found in Table X in Appendix X.}

The lemmatizer has been trained with two configurations. First we used only the training set of the Hungarian UD corpus, then we allowed the tool to learn from the whole transcribed Szeged Corpus (except the sentences overlapping with either our test or development sets).

As for benchmarking other pipelines, we only had to evaluate \emtsv, as the authors of \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} and \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} have already published their tools\textquotesingle \space accuracy on all UD corpora. There are no such published scores available for \emtsv, thus we used the following configuration to produce UD-compatible output for evaluation: \texttt{emToken, emMorph, emLem, emTag, emmorph2ud, emDep, emConll}. While this tool can provide parse trees, their annotation schema is not compatible with that of the Universal Dependencies, hence, its output is not evaluable. We must also note that comparison with the \emtsv{}\textquotesingle s tagger and lemmatizer is not fair, as this tool was trained on a different train-test split which might conflict with ours. (There is a high chance that its training data overlaps with the sentences of our test set.)

\section{Parsing results}

\newlength{\lsz}
\settowidth{\lsz}{Sentence splitting}
\begin{table}
\begin{center}
\begin{tabular}{
  l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lsz}
  >{\centering\arraybackslash}m{\lsz}
}
\toprule
              & Tokenization             & Sentence splitting \\
\midrule
\stanza{}        & 99.87\%                  & 97.00\%            \\
\udpipe{}        & 99.80\%                  & 95.90\%            \\
\emtsv{}         & 99.77\%                  & 98.67\%            \\
\huspacy{} (UD)  & \multirow{2}{*}{99.89\%} & 99.00\%            \\
\huspacy{} (SZC) &                          & 98.01\%            \\  
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Comparison of the tokenization and sentence boundary detection of NLP pipelines measured in F1 scores on the test of the Hungarian UD Corpus}
\label{table:tokens}
\end{center}
\end{table}

Error propagation is a notable phenomenon in language processing applications, thus using highly accurate preprocessing modules such as tokenization and SBD is indispensable. F1 scores in Table \ref{table:tokens}  suggest that tokenization is easily handled by all of the systems making only a few errors, although \huspacy{} is marginally better compared to the rest of the systems. Sentence boundary detection is a more complex task, where language specific knowledge is necessary. This can be either built into the system (as it is the case with \emtsv{}) or learned by a ML model. Numbers show that both approaches can yield high scoring systems.
%\todo{@Zsolti: honnan lett 99\%-os SBD szamunk?}

\newlength{\lm}
\settowidth{\lm}{Morph. acc.}
\begin{table}
\begin{center}
\begin{tabular}{
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
  >{\centering\arraybackslash}m{\lm}
}
\toprule
              & PoS acc. & Morph. acc. & UAS   & LAS   \\
\midrule
\stanza{}        & 96.03\%      & 93.76\%               & 83.62\%   & 78.86\%   \\
\udpipe{} v1     & 90.60\%      & 88.50\%               & 72.80\%   & 67.20\%   \\
\emtsv{}         & 89.19\%      & 89.12\%              & \multicolumn{2}{c}{--\footnote{Although \emtsv can yield parse trees, its annotation schema is not compatible with our test set.}} \\
\huspacy{} (UD)  & 94.87\%      & 89.49\%               & 79.77\%   & 69.96\%   \\
\huspacy{} (SZC) & 96.65\%      & 93.19\%               & 80.09\%   & 71.02\%   \\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Comparison of NLP pipelines in part-of-speech and full morphosyntactic tagging accuracy with attachment scores for dependency parsing on the test set of the Hungarian UD Corpus.}
\label{table:tagging}
\end{center}
\end{table}

Part-of-speech, full morphosyntactic tagging accuracy and attachment scores are presented in Table \ref{table:tagging}. Results show that the overall performance of \huspacy{} and \stanza{} are close to each other. While the latter is better at PoS tagging, the former has significantly higher scores in dependency parsing and predicting morphosyntactic features. It can be seen that the usage of the extra silver standard data yields better performance, both during tagging and dependency parsing for HuSpaCy. \udpipe{} and \emtsv{} have remarkably lower scores compared to our system. While the first can be explained by the underlying simpler model, \emtsv{}\textquotesingle s results are surprising given that it is built upon state-of-the-art morphosyntactic tagging facilities \citep{purepos}. 
%Our assumption is that there could be tag conversion compatibility issues which could cause such low scores. However, this hypothesis needs further investigation.

\newlength{\llem}
\settowidth{\llem}{Accuracy}
\begin{table}
\begin{center}
\begin{tabular}{
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\llem}
}
\toprule
              & Accuracy \\
\midrule
\stanza{}        & 94.25\%    \\
\udpipe{}        & 88.50\%    \\
\udpipe{}         & 94.94\%    \\
\huspacy{} (UD)  & 94.82\%    \\
\huspacy{} (SZC) & 95.43\%    \\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Lemmatization accuracy of NLP pipelines measured on the test set of the Hungarian UD Corpus. \huspacy{} (UD) uses the same setting as its contenders, while \huspacy{} (SZC) uses the whole Szeged Corpus (except the sentences of the test set) for training.}
\label{table:lemma}
\end{center}
\end{table}

Lemmatization results in Table \ref{table:lemma} show that all the systems except \udpipe{} are accurate enough. \huspacy{} trained on the full Szeged Corpus stands out, its score is more than 0.5\% higher than the second best system (\emtsv{}). The best configuration of \huspacy{} system scores more than 0.5\% higher than the one trained solely on the UD dataset. 
%This results confirms the usage of the additional training data.

Resource usage such as memory consumption and processing speed is an important aspect of practical text processing systems, thus we benchmarked\footnote{All experiments were performed on a computer having an Intel Core i7-8750H CPU and 16 GB RAM running Ubuntu Linux 20.04 LTS.} all the NLP pipelines at hand in this respect. In order to have a fair comparison, we configured all of the systems to perform only tokenization, sentence splitting, PoS tagging, lemmatization and dependency parsing. As timing measurements should ignore model loading times, Stanza and \udpipe{} were used by their Python interfaces, while \emtsv{} was utilized using its dockerized version through its REST API. We used the same test set as above to measure throughput and peak memory consumption of the tools.

%\todo{ezeket a szamokat jo lenne jobra rendezni}
\newlength{\lper}
\settowidth{\lper}{Throughput (tokens/sec)}
\begin{table}
\begin{center}
\begin{tabular}{
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lper}
  >{\centering\arraybackslash}m{\lper}
}
\toprule
        & Throughput (tokens/sec) & Memory usage (GB) \\
\midrule
\stanza{}  & 222  & 0.9 \\
\udpipe{}  & 1741 & 0.4 \\
\emtsv{}   & 122  & 3.9 \\
\huspacy{} & 2612 & 2.1 \\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Throughput (measured in tokens/second) and peak memory consumption of benchmarked NLP pipelines.}
\label{table:performance}
\end{center}
\end{table}

Table \ref{table:performance} 
%\todo{Format numbers as it is the NerKor article} 
presents computational efficiency measures of pipelines benchmarked. It can be seen that our system’s throughput is almost 50\% higher than that of the \udpipe{}, while producing significantly better parses. As regards Stanza, there is a huge tradeoff on having the best dependency parser: it is almost 8 times slower than \udpipe{} and more than 10 times slower compared to our system.

Memory consumption of the pipelines are acceptable as all of them could easily fit in a modern computer’s RAM. HuSpaCy has the highest memory usage which is due to the 300-dimensional word vectors shipped with the model. In comparison, Stanza is the only other tool having word embeddings, however its vectors' sizes are limited to 100.

\section{NER experiments and results}

Comparing named entity results is not as straightforward as it is for parsing subtasks. There are multiple evaluation datasets, but there is no consensus between researchers on their usage. NYTK-NerKor (NerKor) \citep{NerKor} is a relatively new corpus consisting of  1 million tokens, while SzegedNER \citep{szeged-ner-corpus} is a 200,000 token subset of the Szeged Corpus.  \citet{nerkor-eval} uses the former dataset to benchmark some of the most popular tools, while previous work mainly rely on the latter one.
 \udpipe{} does not have a NER component, thus we cannot include it in this investigation. As for \emtsv, its entity recognizer was trained using the whole SzegedNER corpus, its comparison against other tools would not be fair.

\huspacy\textquotesingle s entity recognition capabilities are benchmarked in this work on both corpora using the same train-test splits as \citet{szarvas-ner} and \citet{nerkor-eval} suggest. 
As Hungarian entity recognition datasets share the same tagset and rely on similar annotation guides, it is possible to train models using both corpora. In this regards, we follow the work of \citet{nerkor-eval} and evaluate \huspacy{} on the combined corpus as well.
We also include results of previous entity recognition attempts so as to put our results in context. One of the first systems was developed by \cite{szarvas-ner}, which utilizes decision trees for tackling the problem. Next, there is HunTag \citep{huntag, simon-ner}, which is a statistical tagger utilizing a linear model combined with Hidden Markov models. \citet{simon-ner} also showed that it is possible to improve on the F1 score of the base system by incorporating silver standard data. Most recently, \cite{embert} developed an entity recognizer on top of Hungarian BERT models \citep{hubert} achieving state-of-the-art results.

\newlength{\lnersz}
\settowidth{\lnersz}{SzegedNer}

\begin{table}
\begin{center}
\begin{tabular}[t]{
    l<{\hspace{1em}}
  >{\centering\arraybackslash}m{\lnersz}
  >{\centering\arraybackslash}m{\lnersz}
  >{\centering\arraybackslash}m{\lnersz}
}
\toprule
                           & SzegedNer & NerKor & Combined\\
\midrule
\cite{simon-ner}           & 95.06\%   & --      & -- \\
\cite{szarvas-ner}         & 94.77\%\   & --      & -- \\
\embert{}                  & 97.40\%   & 92.09\% & 92.99\% \\
\stanza{}                  & 91.78\%   & 80.53\% & 83.75\% \\
\huspacy{}                 & 95.31\%   & 80.75\% & 83.46\% \\
\bottomrule
\end{tabular}
\vspace{1em}
\caption{Comparison of entity recognition F1 scores on the SzegedNER test set \citep{szarvas-ner}, on the NerKor test set and on the combined test}
\label{table:ner_cmp}
\end{center}
\hspace*{.1\linewidth}
\end{table}

Table \ref{table:ner_cmp} contains a detailed performance comparison of all the entity recognizers mentioned above. Results suggest that the BERT-based model achieves the highest scores on all of the datasets with a large margin. However, these models are also well-known for their enormous computational costs. \huspacy{} is the second best contender on SzegedNER, although its performance is on pair with \stanza{} on other datasets. \embert{}\textquotesingle s results are outstanding when NerKor is involved in the comparison. As \citeauthor{nerkor-eval} concludes these measurements are in accordance with similar English NER benchamrks (cf. \cite{stanza}). Pretrained transformer-based models often yield significantly higher performance scores compared to other sequence tagging approaches due to the underlying attention mechanism and the their model's increased capacity. But there is no free lunch, higher accuracy comes with significantly increased prediction costs. 
%na ezt most jol idehanytam, mindenfele indoklas nelkul... :D

\section{Conclusions}

We presented \huspacy{}, a new industry-ready Hungarian language processing pipeline that is open source and is freely available. While previous approaches have failed to provide a tool which can be easily used to solve practical text processing problems, our system builds on the solid foundations of an industrial NLP framework. We presented how our toolkit utilizes spaCy\textquotesingle s underlying ML models to provide all the basic language analysis facilities with high-enough accuracy and efficient resource usage. We performed various experiments proving that our system has near state-of-the-art accuracy in many text processing tasks while using only moderate computation resources.

As results show, the accuracy of \huspacy\textquotesingle s dependency parser needs further improvements. We plan to tackle this issue by fine-tuning the underlying model through hyperparameter optimization and by incorporating additional silver standard data. Further advancement opportunities lie in fine-tuning the NER model and in using a new neural lemmatizer similar to that of Stanza and TurkuNLP.

In summary, this study described a new freely available tool which is suitable for real-world industrial applications.

% \newlength{\lparams}
% \settowidth{\lparams}{SBD, tagger and parser}
% \begin{table}
% \begin{center}
% \begin{tabular}{
%     l<{\hspace{1em}}
%   >{\centering\arraybackslash}m{\lparams}
%   >{\centering\arraybackslash}m{\lparams}
%   >{\centering\arraybackslash}m{\lparams}
% }
% \toprule
%                       & SBD and tagger                  & SBD, tagger and parser                     & NER \\
% \midrule
% Batch size            & 256        & 256        & 1024       \\
% Encoder width         & 300        & 300        & 128        \\
% Encoder depth         & 4          & 4          & 4          \\
% Encoder window size   & 1          & 1          & 1          \\
% Encoder maxout pieces & 5          & 5          & 3          \\
% Dropout               & 0.1        & 0.1        & 0.1        \\
% Accumulate gradient   & 1          & 1          & 1          \\
% Patience              & 1600       & 1600       & 50000      \\
% Max epochs            & 0          & 0          & 200        \\
% Max steps             & 20000      & 20000      & 0          \\
% Optimizer             & Adam       & Adam       & Adam       \\
% Beta1                 & 0.9        & 0.9        & 0.9        \\
% Beta2                 & 0.999      & 0.999      & 0.999      \\
% L2 is weight decay    & true       & true       & true       \\
% L2                    & 0.01       & 0.01       & 0.01       \\
% Gradient clipping     & 1.0        & 1.0        & 1.0        \\
% Use averages          & false      & false      & false      \\
% Eps                   & $1e^{-9}$  & $1e^{-9}$  & $1e^{-9}$  \\
% Learn rate            & 0.001      & 0.001      & 0.001      \\
% Tag accuracy          & 0.30       & 0.30       & 0.30       \\
% POS accuracy          & 0.15       & 0.15       & 0.15       \\
% Morph accuracy        & 0.15       & 0.15       & 0.15       \\
% UAS                   & 0.15       & 0.15       & 0.15       \\
% LAS                   & 0.15       & 0.15       & 0.15       \\
% Sentences f           & 0.1        & 0.1        & 0          \\
% Entities f            & 0          & 0          & 0.4        \\
% Extra state tokens    &            & false      & false      \\
% Hidden width          &            & 512        & 64         \\
% Maxout pieces         &            & 3          & 2          \\
% Use upper             &            & true       & true       \\
% \bottomrule
% \end{tabular}
% \vspace{1em}
% \caption{tmp}
% \label{table:params}
% \end{center}
% \end{table}

% \section*{Acknowledgements}
% \label{Koszonet}

% ---- Bibliography ----
%
%kulon bibtex fajl hasznalata
\renewcommand\bibname{\refname}
\renewcommand\bibname{References}
%for papers written in English
\bibliographystyle{splncsnat_en}
\bibliography{mszny}

\end{document}

### 2022-01-04 mszny2022/main.tex
Added authors and page numbers

--- 2021-11-23 minta.tex
@@
 
-%\pagestyle{myheadings}
-%\def\leftmark{{\rm XIV. Magyar Sz\'am\'\i t\'og\'epes Nyelv\'eszeti Konferencia}}
-%\def\rightmark{{\rm Szeged, 2018. január 18-19.}}
+\pagestyle{myheadings}
+\def\leftmark{{\rm XVIII. Magyar Sz\'am\'\i t\'og\'epes Nyelv\'eszeti Konferencia}}
+\def\rightmark{{\rm Szeged, 2022. január 27-28.}}
 
 % \setcounter{page}{1}
+% \pagestyle{plain}
 
 \title{\huspacy: an industrial-strength Hungarian natural language processing toolkit}
-%a november 15-ig bekuldott verziobol kerjuk a szerzok adatait kihagyni
-%	\author{Orosz György\inst{1}, Farkas Richárd\inst{1}, Vincze Veronika\inst{1,2}\\
-% \institute{$^1$Szegedi Tudom\'anyegyetem, Informatikai Int\'ezet\break
-%Szeged, \'Arp\'ad t\'er 2.\break
-%$^2$MTA-SZTE Mesters\'eges Intelligencia Kutat\'ocsoport\break
-%Szeged, Tisza Lajos k\"or\'ut 103.\break 
-%\{zsibrita,rfarkas,vinczev\}@inf.u-szeged.hu\break
-%}
-%}
+
+\author{György Orosz\inst{1,2} \and
+    Zsolt Szántó\inst{1,2} \and \break
+    Péter Berkecz\inst{1,2} \and
+    Gergő Szabó\inst{1,2} \and
+    Farkas Richárd\inst{1,2}\\
+    \institute{
+        $^1$Artificial Intelligence National Laboratory \break
+        $^2$Institute of Informatics, University of Szeged\break
+        Szeged Árpád tér 2., H-6720
+    }
+    \email{gyorgy@orosz.link}\break
+    \email{\{szantozs,rfarkas\}@inf.u-szeged.hu}\break
+    \email{\{berkecz.peter,szabogergo.a\}@gmail.com}\break
+}
 
@@
 The underlying models are optimized to be light on memory consumption and CPU usage.
-The presented tool is open source\footnote{The URL is to be shared in the final paper.} and is available under the permissive CC-BY-SA-4.0 license.
+The presented tool is open source\footnote{\url{https://github.com/oroszgy/spacy-hungarian-models}} and is available under the permissive CC-BY-SA-4.0 license.
 Our system is built on top of spaCy’s infrastructure, thus extensive documentation, debugging tools, an ergonomic API and a flourishing ecosystem are already provided.
@@
 \stanza{}        & 94.25\%    \\
-\udpipe{}        & 88.50\%    \\
-\udpipe{}         & 94.94\%    \\
+\udpipe{} v1     & 88.50\%    \\
+\emtsv{}         & 94.94\%    \\
 \huspacy{} (UD)  & 94.82\%    \\
@@
 \bibliographystyle{splncsnat_en}
-\bibliography{mszny}
+\bibliography{references}
 

### 2022-01-05 mszny2022/main.tex
Preprint v1

@@
 
-\author{György Orosz\inst{1,2} \and
-    Zsolt Szántó\inst{1,2} \and \break
-    Péter Berkecz\inst{1,2} \and
-    Gergő Szabó\inst{1,2} \and
-    Farkas Richárd\inst{1,2}\\
+\author{
+György Orosz,
+Zsolt Szántó, \break
+Péter Berkecz,
+Gergő Szabó,
+Richárd Farkas\\
 \institute{
-        $^1$Artificial Intelligence National Laboratory \break
-        $^2$Institute of Informatics, University of Szeged\break
-        Szeged Árpád tér 2., H-6720
+    Institute of Informatics, University of Szeged\break
+    Szeged, Árpád tér 2., H-6720\\
 }
@@
 \email{\{szantozs,rfarkas\}@inf.u-szeged.hu}\break
-    \email{\{berkecz.peter,szabogergo.a\}@gmail.com}\break
 }
@@
 The underlying models are optimized to be light on memory consumption and CPU usage.
-The presented tool is open source\footnote{\url{https://github.com/oroszgy/spacy-hungarian-models}} and is available under the permissive CC-BY-SA-4.0 license.
+The presented tool is open source\footnote{\url{https://github.com/huspacy/huspacy}} and is available under the permissive CC-BY-SA-4.0 license.
 Our system is built on top of spaCy’s infrastructure, thus extensive documentation, debugging tools, an ergonomic API and a flourishing ecosystem are already provided.
@@
 
-% \section*{Acknowledgements}
-% \label{Koszonet}
+\section*{Acknowledgements}
+%\label{Koszonet}
+HuSpaCy research and developments has been financed by the Ministry of Innovation and Technology NRDI Office within the framework of the Artificial Intelligence National Laboratory Program.
 
@@
 \bibliographystyle{splncsnat_en}
-\bibliography{references}
+\bibliography{main}
 

### 2022-01-10 mszny2022/main.tex
Camera ready version

@@
 \restylefloat{figure}
+\usepackage{siunitx}
 \usepackage{tikz-qtree}
@@
 \usepackage{todonotes}
+% \renewcommand\todo{\noop}
 \selectlanguage{english}
@@
 \usepackage{hyperref}
+\hypersetup{plainpages=false,
+            pageanchor=true,
+            colorlinks=true,
+            linkcolor=blue,
+            citecolor=blue,
+            anchorcolor=blue,
+            menucolor=blue,
+            filecolor=blue,
+            urlcolor=blue,
+            linktocpage=true,
+            breaklinks=true,
+            unicode=true}
 \usepackage{soul}
@@
 
-\renewcommand\todo{\noop}
 % \renewcommand\bibname{References}
@@
     Institute of Informatics, University of Szeged\break
-    Szeged, Árpád tér 2., H-6720\\
+    2. Árpád tér, Szeged, Hungary\\
 }
@@
 Industrial text processing applications have to satisfy non-functional software quality requirements, what is more, frameworks supporting multiple languages are more and more favored.
-%have been released in large numbers in recent years, such pipelines also have to satisfy non-functional software quality requirements.  
-%Last, but not least, the international research and industrial community have been experimenting with languages whose language-specific analyzer tools are available in multilingual language processing libraries.
-This paper introduces \huspacy, an industry-ready Hungarian language processing pipeline.
-The presented tool provides components for the most important basic linguistic analysis tasks.
-It is open-source and is available under a permissive license.
-Our system is built upon spaCy’s NLP components which means that it is fast, has a rich ecosystem of NLP applications and extensions, comes with extensive documentation and a well-known API.
-Besides the overview of the underlying models, we also present rigorous evaluation on common benchmark datasets.
-Our experiments confirm that HuSpaCy has high accuracy in all subtasks while maintaining resource-efficient prediction capabilities.
+This paper introduces \huspacy, an industry-ready Hungarian language processing toolkit. 
+The presented tool provides components for the most important basic linguistic analysis tasks. It is open-source and is available under a permissive license. 
+Our system is built upon spaCy’s NLP components resulting in a easily usable, fast yet accurate application.
+Experiments confirm that HuSpaCy has high accuracy while maintaining resource-efficient prediction capabilities. 
 \end{abstract}
@@
 
-Basic natural language processing tasks such as tokenization, sentence splitting, part-of-speech tagging, lemmatization, dependency parsing, named entity recognition and word embedding representations are amongst the most widely studied problems in natural language processing.
+Basic natural language processing tasks such as tokenization, sentence splitting, part-of-speech tagging, lemmatization, dependency parsing and named entity recognition are amongst the most widely studied problems in natural language processing.
 Several text analysis applications have been developed during the last decades for both English and other less-resourced languages such as Hungarian.
-However, a large majority of them solely focus on achieving high accuracy scores on artificial benchmarks and ignore the importance of the underlying model’s practical usability.
+However, a large majority of them solely focus on achieving high scores on artificial benchmarks and ignore the importance of practical usability.
 
@@
 
-Besides the overview of the underlying models, we also  present rigorous evaluation using the Universal grammatical annotations.
-Our experiments confirm that \huspacy has near state-of-the-art accuracy in many subtasks while maintaining resource-efficient prediction capabilities.
+First, we give an overview of the underlying models, then rigorous evaluation is presented using various datasets.
+Finally, experiments are presented confirming that \huspacy{} has high accuracy in many subtasks while maintaining resource-efficient prediction capabilities.
 
@@
 
-In the early 2010s, Universal PoS \citep{udpos} and Universal Dependency \citep{univedepv1} labeling schemata were developed with the goals of ''cross-linguistically consistent treebank annotation for many languages'' and ''facilitating multilingual parser development, cross-lingual learning, and parsing research from a language typology perspective.'' Many language-specific pipelines changed their representations to the universal annotation schema, but most of them stayed in their own software architecture.
+In the early 2010s, Universal PoS \citep{udpos} and Universal Dependency \citep{univedepv1} labeling schemata were developed with the goals of ''cross-linguistically consistent treebank annotation for many languages'' and ''facilitating multilingual parser development, cross-lingual learning, and parsing research from a language typology perspective.'' Many language-specific pipelines changed their representations to these universal annotation schema, but most of them stayed in their own software architecture.
 Industrial NLP applications are frequently multi-lingual, i.e.
@@
 Such systems usually employ a single neural network providing end-to-end NLP solutions without the need for specific pipeline steps.
-Also, several pre-trained multilingual NLMs are becoming available which provide standardised solutions for many languages.
+Also, several pre-trained multilingual NLMs are becoming available which provide standardized solutions for many languages.
 Regardless of the multilinguality and the high accuracy of deep learning solutions, a lot of critiques have been raised by real-world industrial NLP projects recently.
@@
 These drawbacks are questioning the commercial return of the accuracy gain.
-We also note that modern NLP pipelines consist of static word embedding representations and use deep learning for individual pipeline steps as well, hence, the advantage of large neural end-to-end systems
-%\footnote{e.g. BERT models} 
-are very small.
+We also note that modern NLP pipelines consist of static word embedding representations and use deep learning for individual pipeline steps as well, hence, the advantage of large neural end-to-end systems might be very small.
 
@@
 
-Considering decades of experience of practical NLP applications, developing an “industrial strength” text processing system is a challenging task.
+Considering decades of experience of practical NLP applications, developing an “industrial-strength” text processing system is a challenging task.
 First of all, such a tool should tackle the most important text preprocessing tasks including tokenization, sentence splitting, PoS tagging, lemmatization, dependency parsing, named entity recognition and word embedding representation.
@@
 Up until recently, only a few text processing applications were focused on meeting these criteria even for English.
-When spaCy was released in 2015 \citep{spacy}, it was one one of the first tools targeting mainly real-world industrial applications.
+When spaCy was released in 2015 \citep{spacy}, it was one one of the first tools targeting industrial applications mainly.
 Its authors created an unprecedented tool which offers near state-of-the-art accuracy while being an order of magnitude faster than other tools available.
@@
 Although \emtsv{} can yield Universal morphosyntactic annotations through conversion, it is rather inaccurate.
-What is more, it is not designed to efficiently deal with word vectors, therefore no such facility is available in its system.
+What is more, it is not designed to efficiently deal with word vectors, therefore no such facility is available in the system.
 
@@
 \end{center}
+\vspace{-2em}
 \end{table}
@@
 Table \ref{table:systems} summarizes the landscape of the most important Hungarian language processing pipelines and show how they meet the requirements of today’s NLP applications.
+We must note that \emtsv{} does not have any restriction for using it in commercial applications, although some of its most important components have very restrictive licenses (e.g.
+\texttt{emMorph}).
 All in all, we can say that none of them is easily applicable in industrial settings.
 
-We present \huspacy{}, a new industry-ready Hungarian natural language processing toolkit\footnote{We implemented our pipeline in spaCy v3.x.
+We present \huspacy{}, a new industry-ready Hungarian natural language processing toolkit.
+%\footnote{We implemented our pipeline in spaCy v3.x.
 Prior to this work, there were attempts (\url{https://github.com/oroszgy/spacy-hungarian-models}) to support Hungarian in this framework, but none of those releases are compatible with the tool’s current version.}.
@@
 
-\section{Pipeline steps}
+\section{HuSpaCy internals}
 
-This section introduces the NLP algorithms behind \huspacy{}.
-The presented system is built on spaCy’s architecture, therefore we mainly relied on its symbolic and ML-based text processing infrastructure.
-The following paragraphs give only a high-level overview of the internals of the parent system and rather concentrate on the contributions of this work.
+This section introduces the NLP algorithms behind the presented tool.
+As our system is built on spaCy’s architecture, we mainly relied on its symbolic and ML-based text processing infrastructure.
+The following paragraphs give a high-level overview of the framework utilized and also describes the contributions of this work.
 
@@
 
-As regards tokenization, the developed solution uses and extends the underlying system’s algorithm \citep{spacy-tokenization}.
-It works as follows: first the input text is split on whitespaces, then token boundaries are identified by splitting prefixing or suffixing character sequences.
+\huspacy{} builds on spaCy\textquotesingle s \citep{spacy-tokenization} tokenization infrastructure which works as follows: first the input text is split on whitespaces, then token boundaries are identified by splitting prefixing or suffixing character sequences.
 To make this algorithm viable for Hungarian, we extended it with language specific prefix and suffix splitting rules.
-Furthermore, we had to deal with the ambiguity of tokens around the full stop character, thus a decent abbreviation list has been incorporated to increase the module’s accuracy.
-During this process we mostly relied on the tests of \huntoken{} \citep{huntoken}.
+Furthermore, we had to deal with the ambiguity of tokens around full stops, thus an extensive abbreviation list has been incorporated to increase the module\textquotesingle s accuracy.
+During this process we mostly relied on the test cases of \huntoken{} \citep{huntoken} to fine-tune the algorithm.
 
@@
 Sentence boundaries, dependency parse trees, PoS tags, and the corresponding morphosyntactic features are predicted by a multitask deep learning model of the underlying NLP framework.
-SpaCy’s machine learning approach can be summarized as ''embed, encode, attend, predict''  \citep{spacy-neural-model} which our system adapts for its tagging and parsing components.
+SpaCy\textquotesingle s machine learning approach can be summarized as ''embed, encode, attend, predict''  \citep{spacy-neural-model} which our system adapts for its tagging and parsing components.
 
@@
 We use a publicly\footnote{\url{https://github.com/oroszgy/hunlp-resources/releases/tag/webcorpuswiki_word2vec_v0.1}} available 300d word embedding which has been trained on the Hungarian Webcorpus \citep{HunWebCorpus} and a snapshot of the Hungarian Wikipedia with CBOW methodology \citep{word2vec}.
-Task specific word vectors are 256 wide consisting 64 dimensional embeddings of the tokens’ prefixes, suffixes, shapes and the lowercase forms.
+Task specific word vectors are 256 wide consisting 64 dimensional embeddings of the tokens\textquotesingle{} prefixes, suffixes, shapes and the lowercase forms.
 To make such computation efficient, feature hashing is extensively applied to all kinds of input strings.
@@
 During the encoding part, vectors are passed through a four deep stacked CNN encoder \citep{cnn} which uses residual connections and is accompanied with maxout pooling\footnote{The pooling step is considered to be the “attention” mechanism.} \citep{spacy-parser2}.
-The tagger predicts supertags for tokens which are the combinations of the dependency, part-of-speech and morphosyntactic labels.
+%The tagger predicts supertags for tokens which are the combinations of the dependency, part-of-speech and morphosyntactic labels. 
 Efficient prediction is guaranteed by the underlying greedy tagger consisting only of a linear and a softmax layer.
@@
 Finally, sentence boundary recognition is formalized as a sequence tagging problem where tokens are tagged with a binary label indicating the first token of a sentence.
-This component is an integral part of the multitask architecture, thus it also shares its lower layers with the parser and the morphosyntactic tagger.
+This component is an integral part of the multitask architecture, thus it also shares its neural model with the parser and the morphosyntactic tagger.
 
@@
 
-SpaCy’s default lemmatization model is mainly designed for English.
+SpaCy\textquotesingle s default lemmatization model is mainly designed for English.
 It is not suitable for morphologically complex languages such as Hungarian as it only uses lookup tables.
-Hence, we decided to look for a more sophisticated solution, and integrated the \lemmy{} toolkit \citep{lemmy} which is an open-source Python implementation of the CST rule-learning engine \citep{cst-lemmatizer}.
+Hence, we decided to look for a more sophisticated solution and adapted the \lemmy{} toolkit \citep{lemmy} which is an open-source Python implementation of the CST rule-learning engine \citep{cst-lemmatizer}.
 To improve its accuracy we incorporated three minor modifications.
-First, prefixing numbers of numeric tokens are masked to help the generalization of the engine for inflected numbers.
-(For example the masked token of `2021-ben’ becomes `0000-ben’.) Second, we enforce lowercase lemmata of sentence starting tokens if they are not proper nouns.
+First, prefixing numbers of numeric tokens are masked to help the engine in case of inflected numbers.
+(For example the masked token of `2021-ben’ becomes `0000-ben’.) Second, we enforce lowercasing of sentence starting tokens if they are not proper nouns.
 Finally, if there are multiple lemma candidates available for a given (word, tag) pair, we pick the one with the highest frequency on the training dataset.
@@
 SpaCy’s entity recognizer is built on the transition-based parser architecture described in Section \ref{sec:tagging} (similarly to \citet{spacy-ner}). 
-However, there are two key differences compared to the system of \citet{spacy-ner} The first is that the set of possible transition actions reflects the BILOU tagging scheme.
+However, there are two key differences compared to the system of \citet{spacy-ner}. 
+The first is that the set of possible transition actions reflects the BILOU tagging scheme.
 This trick allows the model to have better discrimination ability between different entity classes, furthermore it makes the learning problem easier.
 Second, the state vector computation includes clues not just from the surrounding words but the tokens of previous entities as well. 
-The sequence tagger model uses BILOU tags for encoding entity boundaries and the decoder is based on a simple greedy softmax layer similar to the morphosyntactic tagger.
+The sequence tagger model uses BILOU tags for encoding entity boundaries and the decoder is built on a greedy softmax layer similar to that of the morphosyntactic tagger.
 
-\section{Text parsing experiments}
+\section{Experiments and results}
+\subsection{Text parsing}
+\label{parser-eval}
 
-In order to show \huspacy{} abilities, we performed a series of experiments and compared it with the most popular Hungarian NLP pipelines available.
-For each component, we measured its performance on benchmark datasets along with the model’s throughput and memory consumption.
+In order to benchmark \huspacy{}, we performed a series of experiments comparing its performance with the most popular off-the-shelf pipelines available.
+Evaluation is carried out on the test set of the Hungarian Universal Dependencies Corpus \citep{UniDep} by using the evaluation script of the CoNLL 2018 Shared Task\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.
+
+%It is worth noting that this dataset is a manually transcribed part of the Szeged Corpus.
+
+% / F1-score of each text procesin performance on benchmark datasets along with the model’s throughput and memory consumption.
 As named entity recognition is usually not part of such pipelines, first we focus on evaluating the text parsing task including tokenization, sentence splitting, morphosyntactic tagging, lemmatization and dependency parsing.
 
-We compare our system with the Hungarian specific \emtsv{} pipeline and two other multilingual toolkits, namely \udpipe{} and \stanza{}.
+Three popular text processing tools have been selected for comparison.
+\emtsv{} is a Hungarian specific pipeline integrating state-of-the-art NLP components, \udpipe{} is used as a baseline system in CoNLL competitions, while \stanza{} has high scores on parsing UD corpora.
+All systems are used  as black boxes meaning they have not been retrained or fine-tuned.
+
+% and two other multilingual toolkits, namely \udpipe{} and \stanza{}.
 The latter tools score high on parsing UD corpora and both are off-the-shelf available systems.
@@
 What is more, UD annotations are available only in a rather small subcorpus of it \citep{UniDep}.
-While morphosyntactic labels can be transcribed automatically from the Hungarian-specific formalism to UD with high accuracy, the same is not true for dependency parse trees.
-Hence, there is a notable difference in size of the training data available for tagging and parsing.
+As PoS and morphosyntactic labels can be transcribed automatically from the Hungarian-specific formalism to UD with high accuracy, additional silver standard data can be utilized to train taggers.
 
-To best utilize all possible data sources, we applied a two-step learning strategy
-%\footnote{Our experiments are fully reproducible. Hyperparameters of the models lie in configuration files of the given repository.}
-using \huspacy\textquotesingle s multitask deep learning model.
-In the first step, the tagger and the SBD components are pre-trained on the full-size transcribed Szeged Corpus\footnote{When we refer to the Szeged Corpus as a training set, we mean all the sentences that are not part of the development or the test set of the Universal Dependencies corpus.}.
-This is followed by a fine-tuning step on the gold standard dataset where dependency annotations are also learned.
-A single step model relying solely on the UD data is also involved in our experiments to allow fair comparison with other tools.
+In case of \huspacy{}, we applied a two-step learning strategy
+\footnote{Hyperparameters of the models are available in the tool\textquotesingle s repository (tag \texttt{v0.4.2}) as configuration files.} to best utilize all available training data. 
+In the first step, the tagger and the SBD components are pre-trained on the whole transcribed SZC\footnote{When we refer to the Szeged Corpus as a training set, we mean all the sentences that are not part of the development or the test set of the Universal Dependencies corpus.}.
+This is followed by a fine-tuning step on the gold standard UD dataset where dependency annotations are also learned.
+To allow fair comparison with \stanza{} and \udpipe{}, a single step model relying solely on the UD data is also involved in the evaluation. 
 %%\hl{A summary of hyperparameters can be found in Table X in Appendix X.}
 
-The lemmatizer has been trained with two configurations.
+For similar reasons, the lemmatizer has been trained with two configurations.
 First we used only the training set of the Hungarian UD corpus, then we allowed the tool to learn from the whole transcribed Szeged Corpus (except the sentences overlapping with either our test or development sets).
 
-As for benchmarking other pipelines, we only had to evaluate \emtsv, as the authors of \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} and \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} have already published their tools\textquotesingle \space accuracy on all UD corpora.
-There are no such published scores available for \emtsv, thus we used the following configuration to produce UD-compatible output for evaluation: \texttt{emToken, emMorph, emLem, emTag, emmorph2ud, emDep, emConll}.
-While this tool can provide parse trees, their annotation schema is not compatible with that of the Universal Dependencies, hence, its output is not evaluable.
-We must also note that comparison with the \emtsv{}\textquotesingle s tagger and lemmatizer is not fair, as this tool was trained on a different train-test split which might conflict with ours.
+The authors of \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} and \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} have already published their tools\textquotesingle \space accuracy on the Hungarian UD corpus, however the same is not true for \emtsv{.}
+To evaluate the latter toolkit we used the following (default) configuration to produce an UD-compatible output: \texttt{emToken, emMorph, emLem, emTag, emmorph2ud, emDep, emConll}. 
+While \emtsv{} can provide parse trees, their annotation schema is not compatible with that of the Universal Dependencies, hence, its output is not evaluable. 
+We must also note that comparison with the \emtsv{}\textquotesingle s tagger and lemmatizer might not be fair, as this tool was trained on a different train-test split which might conflict with ours.
 (There is a high chance that its training data overlaps with the sentences of our test set.)
-
-\section{Parsing results}
 
@@
 \emtsv{}         & 99.77\%                  & 98.67\%            \\
-\huspacy{} (UD)  & \multirow{2}{*}{99.89\%} & 99.00\%            \\
-\huspacy{} (SZC) &                          & 98.01\%            \\  
+\huspacy{} (UD)  & \multirow{2}{*}{99.89\%}     & 97.66\%            \\
+\huspacy{} (SZC) &                          & 97.54\%            \\  
 \bottomrule
@@
 \vspace{1em}
-\caption{Comparison of the tokenization and sentence boundary detection of NLP pipelines measured in F1 scores on the test of the Hungarian UD Corpus}
+\caption{Tokenization and sentence boundary detection F1 scores on the test of the Hungarian UD Corpus}
 \label{table:tokens}
 \end{center}
+\vspace{-3em}
 \end{table}
 
-Error propagation is a notable phenomenon in language processing applications, thus using highly accurate preprocessing modules such as tokenization and SBD is indispensable.
-F1 scores in Table \ref{table:tokens}  suggest that tokenization is easily handled by all of the systems making only a few errors, although \huspacy{} is marginally better compared to the rest of the systems.
+% Error propagation is a notable phenomenon in language processing applications, thus using highly accurate preprocessing modules such as tokenization and SBD is indispensable. 
+F1 scores in Table \ref{table:tokens}  suggest that tokenization is easily handled by all of the systems, although \huspacy{} is marginally better compared to the rest of the tools. 
 Sentence boundary detection is a more complex task, where language specific knowledge is necessary.
 This can be either built into the system (as it is the case with \emtsv{}) or learned by a ML model.
-Numbers show that both approaches can yield high scoring systems.
-%\todo{@Zsolti: honnan lett 99\%-os SBD szamunk?}
+Numbers show that both approaches can yield satisfactory SBD components, although the rule-based solution of \emtsv{} stands out followed by the tagging approach of our pipeline.
 
@@
 \udpipe{} v1     & 90.60\%      & 88.50\%               & 72.80\%   & 67.20\%   \\
-\emtsv{}         & 89.19\%      & 89.12\%              & \multicolumn{2}{c}{--\footnote{Although \emtsv can yield parse trees, its annotation schema is not compatible with our test set.}} \\
-\huspacy{} (UD)  & 94.87\%      & 89.49\%               & 79.77\%   & 69.96\%   \\
-\huspacy{} (SZC) & 96.65\%      & 93.19\%               & 80.09\%   & 71.02\%   \\
+\emtsv{}         & 89.19\%      & 89.12\%               & \multicolumn{2}{c}{--} \\
+\huspacy{} (UD)  & 94.70\%      & 89.03\%               & 79.03\%   & 73.17\%   \\
+\huspacy{} (SZC) & 96.58\%      & 93.23\%               & 79.39\%   & 74.22\%   \\
 \bottomrule
@@
 \vspace{1em}
-\caption{Comparison of NLP pipelines in part-of-speech and full morphosyntactic tagging accuracy with attachment scores for dependency parsing on the test set of the Hungarian UD Corpus.}
+\caption{Comparison of  tagging accuracy and attachment scores of the benchmarked pipelines on the test set of the Hungarian UD Corpus.}
 \label{table:tagging}
 \end{center}
+\vspace{-3em}
 \end{table}
 
-Part-of-speech, full morphosyntactic tagging accuracy and attachment scores are presented in Table \ref{table:tagging}.
-Results show that the overall performance of \huspacy{} and \stanza{} are close to each other.
-While the latter is better at PoS tagging, the former has significantly higher scores in dependency parsing and predicting morphosyntactic features.
-It can be seen that the usage of the extra silver standard data yields better performance, both during tagging and dependency parsing for HuSpaCy.
-\udpipe{} and \emtsv{} have remarkably lower scores compared to our system.
-While the first can be explained by the underlying simpler model, \emtsv{}\textquotesingle s results are surprising given that it is built upon state-of-the-art morphosyntactic tagging facilities \citep{purepos}.
+Tagging accuracy and attachment scores are presented in Table \ref{table:tagging}. 
+Results show that \stanza{} is a clear winner in dependency parsing while the PoS tagging score of \huspacy{} (the one using additional training data) is the highest one. 
+It can be seen that the usage of the extra silver standard data yields better performance for our models both during tagging and dependency parsing. 
+\udpipe{} and \emtsv{} have relatively low scores: 
+the results of \udpipe{} are not surprising (cf.
+\cite{conll-2018}), but  \emtsv{}\textquotesingle s scores are unexpected given that it is built upon state-of-the-art morphosyntactic tagging facilities \citep{purepos}.
 %Our assumption is that there could be tag conversion compatibility issues which could cause such low scores. However, this hypothesis needs further investigation.
@@
 \huspacy{} (UD)  & 94.82\%    \\
-\huspacy{} (SZC) & 95.43\%    \\
+\huspacy{} (SZC) & 95.53\%    \\
 \bottomrule
@@
 \caption{Lemmatization accuracy of NLP pipelines measured on the test set of the Hungarian UD Corpus.
-\huspacy{} (UD) uses the same setting as its contenders, while \huspacy{} (SZC) uses the whole Szeged Corpus (except the sentences of the test set) for training.}
+\huspacy{} (UD) uses the same setting as its contenders, while \huspacy{} (SZC) builds on the whole Szeged Corpus for training.}
 \label{table:lemma}
 \end{center}
+\vspace{-3em}
 \end{table}
@@
 \huspacy{} trained on the full Szeged Corpus stands out, its score is more than 0.5\% higher than the second best system (\emtsv{}).
-The best configuration of \huspacy{} system scores more than 0.5\% higher than the one trained solely on the UD dataset.
+The best configuration of \huspacy{} scores more than 0.5\% higher than the one trained solely on the UD dataset.
 %This results confirms the usage of the additional training data.
 
-Resource usage such as memory consumption and processing speed is an important aspect of practical text processing systems, thus we benchmarked\footnote{All experiments were performed on a computer having an Intel Core i7-8750H CPU and 16 GB RAM running Ubuntu Linux 20.04 LTS.} all the NLP pipelines at hand in this respect.
-In order to have a fair comparison, we configured all of the systems to perform only tokenization, sentence splitting, PoS tagging, lemmatization and dependency parsing.
-As timing measurements should ignore model loading times, Stanza and \udpipe{} were used by their Python interfaces, while \emtsv{} was utilized using its dockerized version through its REST API.
-We used the same test set as above to measure throughput and peak memory consumption of the tools.
+\subsection{Named entity recognition}
 
-%\todo{ezeket a szamokat jo lenne jobra rendezni}
-\newlength{\lper}
-\settowidth{\lper}{Throughput (tokens/sec)}
-\begin{table}
-\begin{center}
-\begin{tabular}{
-    l<{\hspace{1em}}
-  >{\centering\arraybackslash}m{\lper}
-  >{\centering\arraybackslash}m{\lper}
-}
-\toprule
-        & Throughput (tokens/sec) & Memory usage (GB) \\
-\midrule
-\stanza{}  & 222  & 0.9 \\
-\udpipe{}  & 1741 & 0.4 \\
-\emtsv{}   & 122  & 3.9 \\
-\huspacy{} & 2612 & 2.1 \\
-\bottomrule
-\end{tabular}
-\vspace{1em}
-\caption{Throughput (measured in tokens/second) and peak memory consumption of benchmarked NLP pipelines.}
-\label{table:performance}
-\end{center}
-\end{table}
-
-Table \ref{table:performance} 
-%\todo{Format numbers as it is the NerKor article} 
-presents computational efficiency measures of pipelines benchmarked.
-It can be seen that our system’s throughput is almost 50\% higher than that of the \udpipe{}, while producing significantly better parses.
-As regards Stanza, there is a huge tradeoff on having the best dependency parser: it is almost 8 times slower than \udpipe{} and more than 10 times slower compared to our system.
-
-Memory consumption of the pipelines are acceptable as all of them could easily fit in a modern computer’s RAM.
-HuSpaCy has the highest memory usage which is due to the 300-dimensional word vectors shipped with the model.
-In comparison, Stanza is the only other tool having word embeddings, however its vectors' sizes are limited to 100.
-
-\section{NER experiments and results}
-
-Comparing named entity results is not as straightforward as it is for parsing subtasks.
+Comparing NER components is not as straightforward as it is for the parsing subtasks.
 There are multiple evaluation datasets, but there is no consensus between researchers on their usage.
@@
 \hspace*{.1\linewidth}
+\vspace{-2em}
 \end{table}
 
-Table \ref{table:ner_cmp} contains a detailed performance comparison of all the entity recognizers mentioned above.
-Results suggest that the BERT-based model achieves the highest scores on all of the datasets with a large margin.
+Table \ref{table:ner_cmp} contains F1 scores of all the entity recognizers mentioned above. 
+It can be seen that the BERT-based model achieves the highest scores on all of the datasets with a large margin.
 However, these models are also well-known for their enormous computational costs.
@@
 \embert{}\textquotesingle s results are outstanding when NerKor is involved in the comparison.
-As \citeauthor{nerkor-eval} concludes these measurements are in accordance with similar English NER benchamrks (cf.
+As \citeauthor{nerkor-eval} concludes these measurements are in accordance with similar English NER benchmarks (cf.
 \cite{stanza}).
-Pretrained transformer-based models often yield significantly higher performance scores compared to other sequence tagging approaches due to the underlying attention mechanism and the their model's increased capacity.
+Pretrained transformer-based models often yield significantly higher performance scores compared to other sequence tagging approaches due to the underlying attention mechanism and the their model\textquotesingle s increased capacity.
 But there is no free lunch, higher accuracy comes with significantly increased prediction costs.
-%na ezt most jol idehanytam, mindenfele indoklas nelkul... :D
+
+The final model of \huspacy{} builds on the weights of a pretrained neural tagger (using the strategy described in Section \ref{parser-eval}) yielding 84.56\% F1 on the combined dataset.
+This result is a significant improvement compared to \stanza \textquotesingle s score and also confirms the usefulness of additional silver standard training data usage for spaCy\textquotesingle s multitask neural model.
+
+\subsection{Resource usage}
+Resource usage such as memory consumption and processing speed is an important aspect of practical text processing systems, thus we benchmarked\footnote{All experiments were performed on a computer having an Intel Core i7-8750H CPU and 16 GB RAM running Ubuntu Linux 20.04 LTS.} text parsing pipelines (cf.
+Section \ref{parser-eval}) in this respect.
+In order to have a fair comparison, we configured all systems to perform only tokenization, sentence splitting, PoS tagging, lemmatization and dependency parsing.
+As timing measurements should ignore model loading times, Stanza and \udpipe{} were used by their Python interfaces, while \emtsv{} was utilized through its REST API.
+We used the UD test set to measure throughput and peak memory consumption.
+
+\newlength{\lper}
+\settowidth{\lper}{Throughput (tokens/sec)}
+\begin{table}
+\begin{center}
+\begin{tabular}{
+    l<{\hspace{1em}}
+  r>{\centering\arraybackslash}m{\lper}
+  r>{\centering\arraybackslash}m{\lper}
+}
+\toprule
+        & Throughput (tokens/sec) & Memory usage (GB) \\
+\midrule
+\stanza{}  & \num{222}  & \num{0.9} \\
+\udpipe{}  & \num{1741} & \num{0.4} \\
+\emtsv{}   & \num{122}  & \num{3.9} \\
+\huspacy{} & \num{2612} & \num{2.1} \\
+\bottomrule
+\end{tabular}
+\vspace{1em}
+\caption{Throughput (measured in tokens/second) and peak memory consumption of benchmarked NLP pipelines.}
+\label{table:performance}
+\end{center}
+\vspace{-2em}
+\end{table}
+
+Table \ref{table:performance} 
+presents computational efficiency measures suggesting that our system has the highest throughput amongst all the tools.
+\huspacy{} is almost 50\% faster than \udpipe{}, while producing significantly better parses.
+As regards \stanza{}, there is a huge tradeoff on having the best dependency parser: it is almost 8 times slower than \udpipe{} and more than 10 times slower compared to our system.
+
+Memory consumption of the pipelines are acceptable as all of them could easily fit in a modern computer’s RAM.
+Our tool has the highest memory usage which is due to its 300-dimensional word vectors.
+In comparison, \stanza{} is the only other tool having word embeddings, but its vectors\textquotesingle{} sizes are limited to 100d.
 
@@
 While previous approaches have failed to provide a tool which can be easily used to solve practical text processing problems, our system builds on the solid foundations of an industrial NLP framework.
-We presented how our toolkit utilizes spaCy\textquotesingle s underlying ML models to provide all the basic language analysis facilities with high-enough accuracy and efficient resource usage.
-We performed various experiments proving that our system has near state-of-the-art accuracy in many text processing tasks while using only moderate computation resources.
+We presented how our toolkit utilizes spaCy\textquotesingle s underlying ML models to provide all the basic language analysis components.
+We performed various experiments proving that our system has high accuracy in many text processing tasks while using only moderate computation resources.
 
 As results show, the accuracy of \huspacy\textquotesingle s dependency parser needs further improvements. 
-We plan to tackle this issue by fine-tuning the underlying model through hyperparameter optimization and by incorporating additional silver standard data.
-Further advancement opportunities lie in fine-tuning the NER model and in using a new neural lemmatizer similar to that of Stanza and TurkuNLP.
+% We plan to tackle this issue by fine-tuning the underlying model through hyperparameter optimization and by incorporating additional silver standard data. 
+Further advancement opportunities lie in fine-tuning the NER model and in using a new neural lemmatizer.
 
@@
 %\label{Koszonet}
-HuSpaCy research and developments has been financed by the Ministry of Innovation and Technology NRDI Office within the framework of the Artificial Intelligence National Laboratory Program.
+The authors would like to thank Dávid Nemeskey and Dániel Lévai for their help in benchmarking \embert{} and \stanza{}.
+HuSpaCy research and development is funded by the Ministry of Innovation and Technology NRDI Office within the framework of the Artificial Intelligence National Laboratory Program.
 

### 2023-02-02 mszny2023/hybrid-lemmatization-in-huspacy/main.tex

\documentclass{llncs}

\usepackage[a4paper,includeheadfoot,top=1.65in,bottom=1.65in,left=1.73in,hcentering]{geometry}
\usepackage[pdftex]{graphicx}
\usepackage[sectionbib]{natbib}

\usepackage[utf8]{inputenc}
\usepackage{t1enc}
\usepackage{tikz-dependency}
\usepackage{qtree}
\usepackage{float}
\restylefloat{figure}
\usepackage{siunitx}
\usepackage{tikz-qtree}
\usepackage[english]{babel}
\usepackage{todonotes}
% \renewcommand\todo{\noop}
\selectlanguage{english}

\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{plainpages=false,
            pageanchor=true,
            colorlinks=true,
            linkcolor=blue,
            citecolor=blue,
            anchorcolor=blue,
            menucolor=blue,
            filecolor=blue,
            urlcolor=blue,
            linktocpage=true,
            breaklinks=true,
            unicode=true}
\usepackage{soul}

\newcommand{\embert}{\texttt{emBERT}}
\newcommand{\emtsv}{\texttt{emtsv}}
\newcommand{\floret}{\texttt{floret}}
\newcommand{\fasttext}{\texttt{fastText}}
\newcommand{\hubert}{\texttt{huBERT}}
\newcommand{\xlmroberta}{\texttt{XLM-Roberta}}
\newcommand{\magyarlanc}{\texttt{magyarlanc}}
\newcommand{\spacy}{spaCy}
\newcommand{\udpipe}{UDPipe}
\newcommand{\stanza}{Stanza}
\newcommand{\huspacy}{HuSpaCy}
\newcommand{\huntoken}{HunToken}
\newcommand{\lemmy}{Lemmy}

% \renewcommand\bibname{References}

\begin{document}

% \pagestyle{myheadings}
% \def\leftmark{{\rm XIX. Magyar Sz\'am\'\i t\'og\'epes Nyelv\'eszeti Konferencia}}
% \def\rightmark{{\rm Szeged, 2023. január 26-27.}}

% \setcounter{page}{1}
% \pagestyle{plain}

\title{Hybrid lemmatization in HuSpaCy}

\author{Péter Berkecz, György Orosz, Zsolt Szántó,\\
Gergő Szabó, Richárd Farkas
 \institute{Institute of Informatics, University of Szeged\break
   2. Árpád tér, Szeged, Hungary\break

\{berkecz,szantozs,gszabo,rfarkas\}@inf.u-szeged.hu\break
gyorgy@orosz.link\break
}
}

\maketitle

\begin{abstract}
Lemmatization is still not a trivial task for morphologically rich languages. Previous studies showed that hybrid architectures usually work better for these languages and can yield great results. This paper presents a hybrid lemmatizer utilizing both a neural model, dictionaries and hand-crafted rules. We introduce a hybrid architecture along with empirical results on a widely used Hungarian dataset. The presented methods are published as three HuSpaCy models.
\end{abstract}

\section{Introduction}

Lemmatization is the process where the lemma - the dictionary form of a word - needs to be computed from an inflected word form. This can happen in various ways. One approach is that we are deriving the possible lemmata of a token using a morphological analyzer which relies on hand-crafted morphological rules then a statistical model is applied to disambiguate using the context. 
On the other hand, end-to-end statistical lemmatizers are using a training corpus exclusively to learn the transformation rules and the context disambiguation in a single model. Hence, they can track the language evolution and fit domain variations if a training dataset is available.

Several end-to-end statistical lemmatizers have been proposed \citep{muller-etal-2015-joint, udpipe} however, based on our experiments on the Hungarian subset of Universal Dependencies \citep{nivre2017universal}, they were not accurate enough on their own. Studies \citep{purepos,boudchiche2019hybrid} showed, that for morphologically rich languages hybrid architectures tend to perform better.

In this paper, we propose the following three-step hybrid architecture: 
\begin{enumerate}
    \item a dictionary-based first step for unambiguous cases, where we just replace the token with the lemma if the traits are the same,
    \item end-to-end statistical lemmatizer for the highest accuracy in ambiguous cases
    \item hand-made rules for special cases where the statistical approach did not see enough examples.
\end{enumerate}

For end-to-end lemmatization, we investigated two algorithms that learn rules from training data, the Lemmy \citep{lemmy} and the Edit-tree Lemmatizer \citep{spacy-edit-tree}. Both of them use neural networks to find the best set of rules for a given word in a given context. We examined various word embedding and transformer-based networks to find the best-performing end-to-end lemmatizer for Hungarian.

Our tool is built on spaCy's foundation and is both versatile and easy to use in a multi-language environment. It is conveniently released under a Creative Commons BY-SA 4.0 license, making it an attractive option for organizations looking to incorporate natural language processing capabilities into their systems without any licensing issues.

Our hybrid lemmatizer is available in the open-source HuSpaCy toolkit, in the \texttt{medium}, \texttt{large}, and \texttt{transformer} models since version 3.4.

\section{Related Work}

\subsection{Lemmatizing in spaCy}

spaCy \citep{spacy} is an industrial-strength natural language processing pipeline, which already supports lemmatization for nineteen languages. HuSpaCy \citep{HuSpaCy:2021} is a Python package supporting Hungarian models for spaCy. Three different models are available for the tool, with two different embedding types. The \texttt{medium} and \texttt{large} model uses static word vectors, while the \texttt{transformer} model relies on \hubert{} \citep{hubert}, a fine-tuned BERT \citep{https://doi.org/10.48550/arxiv.1810.04805} model for Hungarian. Up until now, HuSpaCy utilized a simple rule-learning algorithm it used a lemmatizer called Lemmy.

There are three different built-in lemmatizers available in spaCy. The first is the ruled-based one. It uses hand-crafted rules and part-of-speech tags to make a lemma from a token. Four official spaCy models use this, English, French, Spanish, and Macedonian.

The second type is a dictionary-based lemmatizer. It stores the tokens with their PoS tags, and it loads a lemma for a token during prediction. Only one official spaCy model uses this, the Catalan one.

The third and newest one is the Edit-tree Lemmatizer, which is an end-to-end neural lemmatizer. Fourteen official spaCy models are using this, for example, Greek, Polish, Danish, German, and Finnish.

\subsection{Hungarian Lemmatization Approaches}

In Hungarian, there are several approaches to lemmatizing texts, for example, \magyarlanc{}  \citep{magyaralanc} and \emtsv{} \citep{emtsv1, emtsv2, emtsv3}. Both of them use symbolic methods to make lemma candidates and disambiguate between them based on the context of the word in another step.

The morphological parser in \magyarlanc{} generates lemma candidates based on KR \citep{tron2006annotation} grammatical rules and then uses the PurePOS part-of-speech tagger \citep{purepos} to select the correct one according to the sentence context.

\emtsv{} is the result of a collective effort to integrate multiple existing NLP tasks into one application. For lemmatization, it uses the emMorph \citep{novak2016new,novak2014new} morphological analyzer to create a lemma, part-of-speech tag, and morphological description candidates, and the emTag (PurePOS) tagger to disambiguate between them. During prediction, it uses Hungarian rules \citep{novak2015model} for HFST \citep{hfst}, then from the pool of lemma candidates, it selects the most probable one using the PurePOS tagger. We must note that emMorph has a very restrictive license, thus it cannot be used freely in a commercial application.

The lemmatizer of Stanza \citep{stanza} combines a dictionary-based lemmatizer with a neural Seq2Seq lemmatizer. The dictionary  is being used as a cache to predict lemmata quickly because the Seq2Seq lemmatizer is typically slower, this is also compiled from the training dataset.

UDPipe \citep{udpipe} handles lemmatization as a sub-task of tagging. The system generates a rule from tokens and their lemmata based on the longest common subsequence of two words and then selects them as a classification task.

\section{Methods}

Various studies \citep{HuSpaCy:2021, cst-lemmatizer} have already shown how the different types of lemmatizers perform individually. In some cases of agglutinating languages, however, the best solution was usually achieved with a combination of these \citep{orosz2015hybrid}. In the following section, we present the methods we have selected for our architecture.

\subsection{Lemmy}

Lemmy is an end-to-end statistical lemmatizer. It is an open-source Python implementation of the CST algorithm \citep{cst-lemmatizer}. This method automatically learns rules from the training data by detecting prefixes and suffixes to be transformed, so that the resulting token can be used to generate the lemma. First, it tries to lemmatize based on the existing rules, if this fails, it creates a new rule using their longest common subsequence and their differences. It creates a rule based on the token and its lemma’s longest common subsequence, and stores this and the token’s PoS-tag with the rule occurrence count. It had a few modifications over the original project to improve its accuracy. During prediction, it uses this frequency information and the predicted PoS-tag to select a rule, then applies it to the token to compute the lemma.

\subsection{Edit-tree Lemmatizer}

The Edit-tree Lemmatizer is another end-to-end statistical lemmatizer that exploits neural language models. It is a spaCy implementation of the Lemming lemmatizer \citep{muller-etal-2015-joint}, which learns rules (called edit-trees) from the training data to transform a word into its lemma.

For example, this system builds an edit-tree for the word \textit{leghosszabb} (longest) this way:

\begin{enumerate}
    \item Search for the longest common subsequence (LCS) between the token and its lemma: \textit{leghosszabb} - \textit{hosszú} (long)
    \item Based on this, it splits into 3 parts, prefix - LCS - suffix
    \item Finds the necessary transformations for the prefix and suffix to transform the inflected form into the lemma
    \begin{enumerate}
        \item In the prefix, we change \textit{leg-} into an empty string ($\epsilon$)
        \item In the suffix, we change \textit{-abb} into \textit{-ú}
    \end{enumerate}
\end{enumerate}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\textwidth]{edit-tree-fig.pdf}
  \caption{Edit-tree for the word \textit{leghosszabb}}
  \label{fig:edit-tree}
\end{figure}

When predicting, a neural network is used as a classifier to select the right edit-tree. In spaCy we can use multi-task learning and this can be inserted as a task, thus being able to access hidden representations in the underlying neural network. This way, for example, the tagger and the lemmatizer can work together in an even closer relationship, not just on the few tags that the tagger predicts.

\subsection{Hungarian Language Models Used} \label{neuralnetbehind}

The architecture of HuSpaCy \citep{HuSpaCy:2021} consisted of a Word2Vec \citep{word2vec} static word embedding and CNN layers. The disadvantage of Word2Vec is that it only knows the words it saw during training and does not use the context information of the tokens.

To overcome these drawbacks, we experimented with word embeddings containing subword information and context-aware transformer networks.

Static word embeddings containing subword information (e.g. \fasttext{} \citep{bojanowski2017enriching}) learn not just a single word during training, but several small parts of a word. This way, because a token is a sum of its character \textit{n-gram} vectors, it performs better for words outside the vocabulary (OOV), be it a truly unknown word or a typo. It has the advantage of having a small storage footprint due to the smaller vocabulary size and because of that, it is quite fast when evaluated on the CPU. In our experiments, we have utilized \floret{} \citep{floret} which is the spaCy fork of the \fasttext. We trained two models on the Hungarian WebCorpus 2.0 \citep{Nemeskey:2020}, a 100-dimensional and a 300-dimensional one. The advantage of them is that they include character n-grams, and due to the hashing trick of \floret, their storage size is much smaller than the original word embeddings.

On the other hand, transformer models (e.g. \texttt{BERT}) are able to take the full sentence context of the word into consideration when predicting, and thus can give a more accurate representation and achieve a high level of accuracy in HuSpaCy \citep{gergo}. Their disadvantages are their large storage size and relatively slow CPU evaluation, but with GPU it can be fast.

\subsection{Lookup Lemmatizer}

Dictionary-based lemmatizers only perform a simple replacement operation. They “learn” to map tokens to lemmata from the training dataset by memorizing a word form with their morphosyntactic information associated with lemmata. If more than one lemma is associated with the same word with the same morphosyntactic labels, the most frequent one is learned. In prediction time, they simply return the matching lemmata if any. The disadvantage is that it can only lemmatize tokens that it has already seen.

\subsection{Postprocessing Rules} \label{rules}

There may be cases where the lemmatizer makes mistakes and these cases can be easily corrected by rules. These special hand-crafted rules are most often used when something has to be cut off the end of a token to get the correct lemma. You could actually call it the little brother of stemming. Beyond suffix cutting, these rules are also useful in certain cases for agglutinating languages.

We noticed that in a few rare cases where there are exclamation and question marks in a token they were left in a lemma, thus we’ve built a rule to remove them. Also, there were some numbers and dates that needed to be fixed. For example, for the token \textit{4-6-os} the lemmatizers usually made \textit{4-6-} as a lemma, but \textit{4-6} would be the correct one. Because the predicted lemmata had unrecognizable patterns, this rule used the original token to remove any suffixes after a number or a date.

\section{Results}

\subsection{Experimental Setup}

To ensure comparability, we have used the most commonly available tools for training and testing. The evaluations were carried out on the test set of the Hungarian Universal Dependencies (UD) \citep{nivre2017universal} corpus using the CoNLL 2018 Shared Task \citep{conll-2018} evaluator script\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}}.

For training, we used parts of the Szeged Korpusz \citep{szegedcorpus} that are not included in any part of the UD corpus. As for the comparison, we chose the most popular open-source tools and used them as they were published for reproducibility.

\subsection{Experiments}

Lemmy was used in HuSpaCy, but it only relied on the part-of-speech information and did not have any other information about the token. The Edit-tree Lemmatizer was released in spaCy 3.3. Studies \citep{spacy-edit-tree} showed that it brings significant improvements for various languages, hence we investigated its usage for Hungarian (cf. Table \ref{table:baseline}).

\newlength{\ltbaseline}
\settowidth{\ltbaseline}{Edit-tree Lemmatizer}
\begin{table}
	\begin{center}
		\begin{tabular}{
			>{\centering\arraybackslash}m{\ltbaseline}
			>{\centering\arraybackslash}m{\ltbaseline}
			}
			\toprule
			                  Lemmy & Edit-tree Lemmatizer \\
			\midrule
                     95.53\% & 95.90\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Results of the previous Lemmy lemmatizer and the Edit-tree Lemmatizer with the default configurations. We consider this as the baseline.}
		\label{table:baseline}
	\end{center}
	\vspace{-3em}
\end{table}

As we have shown in subsection \ref{neuralnetbehind}, the underlying neural language model can be simply replaced by another spaCy-compatible module. The Word2Vec word embeddings previously shipped with HuSpacy models were replaced by a \floret{} \citep{floret} architecture. This not only improves our results but also reduces the storage requirements of the whole model.

We experimented with pretrained transformer language models as well. Based on our experiments (cf. Table \ref{table:new-models}), \hubert{} \citep{hubert} gave a good accuracy-to-speed ratio, but with an \xlmroberta\texttt{-Large} \citep{https://doi.org/10.48550/arxiv.1911.02116} model we were able to achieve even higher accuracy at the expense of throughput.

\newlength{\ltnew}
\settowidth{\ltnew}{\floret (100d)}
\newlength{\ltroberta}
\settowidth{\ltroberta}{\xlmroberta}
\begin{table}
	\begin{center}
		\begin{tabular}{
			>{\centering\arraybackslash}m{\ltnew}
			>{\centering\arraybackslash}m{\ltnew}
			>{\centering\arraybackslash}m{\ltnew}
			>{\centering\arraybackslash}m{\ltroberta}
			}
			\toprule
			      \floret (100d)& \floret (300d) & \hubert & \xlmroberta \\
			\midrule
                     96.56\% & 96.76\% & 98.53\% & 98.89\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Results of the Edit-tree Lemmatizer with different language models.}
		\label{table:new-models}
	\end{center}
	\vspace{-3em}
\end{table}

The Edit-tree Lemmatizer has an important hyperparameter, which controls the number of edit-trees to try before giving up and choosing another one of the token's properties as a lemma. It defaults to 1, but increasing this value improved the accuracy. By adjusting this, the model can access more edit-tree candidates during prediction because the possible edit-tree is chosen from all of them. Increasing it to 3 improves the result (cf. Table \ref{table:topk}) even though the pipeline is slower, but it's a worthy compromise. A higher \texttt{top\_k} value makes the algorithm a bit slower (cf. Table \ref{table:topk-speed}) and has almost no effect on the results. Because of this, we've choosen 3 as our \texttt{top\_k} value for the next experiments.

\newlength{\lttopk}
\settowidth{\lttopk}{\floret (100d)}
\begin{table}
	\begin{center}
		\begin{tabular}{
			>{\centering\arraybackslash}l{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\ltroberta}
			}
			\toprule
			       & \floret (100d)& \floret (300d) & \hubert & \xlmroberta \\
			\midrule
                     \texttt{top\_k} = 1 & 96.56\% & 96.76\% & 98.53\% & 98.89\% \\
                     \texttt{top\_k} = 3 & 96.87\% & 97.01\% & 98.63\% & 98.93\% \\
                     \texttt{top\_k} = 8 & 96.84\% & 97.09\% & 98.61\% & 98.83\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Lemma accuracies of the Edit-tree Lemmatizer with increased \texttt{top\_k} parameters.}
		\label{table:topk}
	\end{center}
	\vspace{-3em}
\end{table}

\begin{table}
	\begin{center}
            \vspace{-3em}
		\begin{tabular}{
			>{\centering\arraybackslash}l{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\lttopk}
			>{\centering\arraybackslash}m{\ltroberta}
			}
			\toprule
			       & \floret (100d)& \floret (300d) & \hubert & \xlmroberta \\
			\midrule
                     \texttt{top\_k} = 1 & 7739 & 7255 & 3349 & 2429 \\
                     \texttt{top\_k} = 3 & 6740 & 6873 & 3166 & 2358 \\
                     \texttt{top\_k} = 8 & 6655 & 6571 & 3146 & 2347 \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Throughput (token/s) of the Edit-tree Lemmatizer with increased \texttt{top\_k} parameters.}
		\label{table:topk-speed}
	\end{center}
	\vspace{-3em}
\end{table}

In Lemmy, it was already noticeable that the casing of some words was wrong, this problem was also apparent here. So the same rule had to be built into the system (cf. Table \ref{table:etl-casing}). If the word is a proper noun or the first token of a sentence, it can start with a capital letter, in all other cases, the word is lowercase. We built this in so that the algorithm can be aware of these casings during training so that it can apply them later during prediction.

\newlength{\ltetlcasing}
\settowidth{\ltetlcasing}{\floret (100d)}
\begin{table}
	\begin{center}
		\begin{tabular}{
			>{\centering\arraybackslash}l{\ltlookup}
			>{\centering\arraybackslash}m{\ltetlcasing}
			>{\centering\arraybackslash}m{\ltetlcasing}
			>{\centering\arraybackslash}m{\ltetlcasing}
			>{\centering\arraybackslash}m{\ltroberta}
			}
			\toprule
			     & \floret (100d)& \floret (300d) & \hubert & \xlmroberta \\
			\midrule
                   + Casing & 96.29\% & 97.23\% & 98.63\% & 98.85\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Results of the Edit-tree Lemmatizer taking casing into account.}
		\label{table:etl-casing}
	\end{center}
	\vspace{-3em}
\end{table}

\subsubsection{Hybrid Lemmatizer Results}

Since the Edit-tree Lemmatizer doesn't give any prediction sometimes, we tested how a dictionary-based lemmatizer performs as a supplement. This became the second part of our architecture, but it became a pipeline step before the Edit-tree Lemmatizer in the whole spaCy pipeline.

We had two experiments with this setup, the first was about pairing the training data to the lemmata based on the token and PoS tag (cf. Table \ref{table:lookup-rules}). However, this led to many incorrect predictions. Finally, to solve this, we paired the training data to the lemmata to be trained based on the token, the PoS-tag, and the morphological tags. During training and prediction, the number-like tokens are masked, resulting in fewer pairs, but this will match more tokens containing numbers. It can generalize better with this because for example the \textit{1000-ben} will be \textit{0000-ben} and it could be applied to \textit{3000-ben} which may not have been included in the training set.

In a couple of cases, even combined -- the dictionary-based lemmatizer with the Edit-tree Lemmatizer -- they still made some errors, and we used post-processing rules to try to correct the lemma in the ways that we mention in subsection \ref{rules}. Upon implementation of these rules, the accuracies demonstrated a slight decrease, however, the overall quality of the predicted lemmata  showed noticeable improvement.

\newlength{\ltlookup}
\settowidth{\ltlookup}{\floret (100d)}
\begin{table}
	\begin{center}
		\begin{tabular}{
			>{\centering\arraybackslash}l{\ltlookup}
			>{\centering\arraybackslash}m{\ltlookup}
			>{\centering\arraybackslash}m{\ltlookup}
			>{\centering\arraybackslash}m{\ltlookup}
			>{\centering\arraybackslash}m{\ltroberta}
			}
			\toprule
			       & \floret (100d)& \floret (300d) & \hubert & \xlmroberta \\
			\midrule
                     + Lookup & 97.19\% & 97.48\% & 98.68\% & 98.94\% \\
                     + Rules & 97.17\% & 97.46\% & 98.68\% & 98.94\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Results of the hybrid architecture. The rows are an addition to the existing system. It can be observed that the transformer-based models were not really helped by the hybrid additions, but the floret-based models were. This is due to the fact that the larger models are more adept at selecting the edit-trees.}
		\label{table:lookup-rules}
	\end{center}
	\vspace{-3em}
\end{table}

\subsection{Final Results}

In order to compare our architecture's results with already existing systems, we chose three popular systems that have built-in lemmatizers. \emtsv{} is a Hungarian pipeline of state-of-the-art NLP components, \udpipe{} is commonly used as a baseline in CoNLL competitions, and \stanza{} has relatively high scores on UD test sets. It is important to point out that all of them were used off-the-shelf, i.e. not re-trained or fine-tuned.

The evaluation was performed on the Hungarian Universal Dependencies test set. The \udpipe{} and \stanza{} authors have made the results of their systems available, but the \emtsv{} authors have not, so we used one of the default configurations to get UD-compatible output and then evaluate it. It is important to point out that comparing \emtsv's lemmatizer is probably not fair out-of-the-box, because the tool was trained on a different train-test split, which may conflict with ours. (There is a high probability that their training set contains sentences in our test set.)

\newlength{\ltcompare}
\settowidth{\ltcompare}{Accuracy}
\begin{table}
	\begin{center}
		\begin{tabular}{
			l<{\hspace{1em}}
			>{\centering\arraybackslash}m{\ltcompare}
			}
			\toprule
			                 & Accuracy \\
   \midrule
                Word2Vec \textit{baseline} & 95.90\% \\
   \midrule
			HuSpaCy + \floret{} (100d) & 97.17\%  \\
			HuSpaCy + \floret{} (300d) & 97.46\%  \\
			HuSpaCy + \hubert{}        & 98.68\%  \\
			HuSpaCy + \xlmroberta{}    & 98.94\%  \\
   \midrule
                \emtsv{}  & 96.16\% \\
                \udpipe{} & 88.50\% \\
                \stanza{} & 94.19\% \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Lemmatization accuracies of the different NLP pipelines.}
		\label{table:compare}
	\end{center}
	\vspace{-3em}
\end{table}

In Table \ref{table:compare} one can see that all systems are quite accurate except \udpipe{}. Our hybrid lemmatizer consistently outperforms other systems, demonstrating a marked improvement in performance.

\subsection{Resource Usage}

It is also important to mention the resource usage of our models. These models contain a complete language pipeline, i.e. a single model of a system. In the case of HuSpaCy, these models include tagger, morphologizer, lemmatizer, parser, and named entity recognition.

\newlength{\ltres}
\settowidth{\ltres}{Storage Size}
\begin{table}
	\begin{center}
		\begin{tabular}{
			l<{\hspace{1em}}
			>{\centering\arraybackslash}m{\ltres}
			>{\centering\arraybackslash}m{\ltres}
			>{\centering\arraybackslash}m{\ltres}
			>{\centering\arraybackslash}m{\ltres}
			}
			\toprule
			                 & Storage Size  &  Throughput (tokens/sec) CPU &  Throughput (tokens/sec) GPU & Memory Usage (GiB)\\
			\midrule
			Word2Vec \textit{baseline} & 1.6 GB &  854 & 7059 &  4.61 \\
   \midrule
			HuSpaCy + \floret{} (100d) & 125 MB & 1862 & 5903 &  2.35 \\
			HuSpaCy + \floret{} (300d) & 455 MB & 1186 & 6074 &  3.33 \\
			HuSpaCy + \hubert{}        & 1.8 GB &  242 & 3055 &  4.84 \\
			HuSpaCy + \xlmroberta{}    & 9.1 GB &   78 & 2186 & 17.87 \\
   \midrule
                \emtsv{}  & 3.6 GB &  116 &   - & 3.91 \\
                \udpipe{} &   5 MB & 3175 &   - & 1.38 \\
                \stanza{} & 246 MB &   30 & 395 & 5.31 \\
			\bottomrule
		\end{tabular}
		\vspace{1em}
		\caption{Resource usages of the different models. The storage sizes contain a full natural language processing pipeline, not just the lemmatizer. During measurement, we've tried to run every benchmark three times, and take the best result with each tool. For each tool, we've tried to use only the parts that are needed to the lemmatizer, in the case of our models this means that we've run them without named entity recognition. We measured the throughput using an AMD EPYC 7F72 CPU and NVIDIA A100 (40GB) GPU.}
		\label{table:resource}
	\end{center}
	\vspace{-3em}
\end{table}

Table \ref{table:resource} shows that the results do not really correlate with the storage requirements of a model. This is mostly due to the language model, as can be seen in the case of the Word2Vec model, which is 1.6GB but is outperformed by a 125MB model in accuracy. However, the two models use completely different word embeddings, while Word2Vec uses a continuous bag-of-words model, \floret{} uses CBOW with subword information but also hashes words so it is a bit slower.

For both \xlmroberta-based and \hubert-based, we can see that the 5x storage requirement is almost unnecessary, with the result being only a few tenths of a percentage point better.

We must note that at \emtsv{} it is the size of a Docker image which is a recommended way to use the tool, thus it may contain parts that are not part of the original \emtsv{} pipeline.

\section{Conclusions}

We presented a hybrid architecture, which consists of three parts utilizing the advantages of neural language model-based statistical lemmatizers and dictionaries along with hand-crafted rules. We have shown the application of hand-crafted rules and morphological information-based dictionaries can improve end-to-end statistical lemmatizers. We examined different language models and we found the \xlmroberta\texttt{-large} model achieves the best results, but the five times smaller \hubert{} reached comparable results too.  We performed multiple experiments to extensively compare our tool with other systems. As the results show, our lemmatizer outperforms other models and the storage requirement is smaller with the smaller \floret{} model than some of them.

Altogether, this paper has presented a new hybrid lemmatizer architecture that is freely available in three models of HuSpaCy. We made it open-source, under a permissive license.

\section*{Acknowledgements}
%\label{Koszonet}
The authors would like to thank Dávid Nemeskey and Dániel Lévai for their help in benchmarking \embert{} and \stanza{}.
HuSpaCy research and development is supported by the European Union project RRF-2.3.1-21-2022-00004 within the framework of the Artificial Intelligence National Laboratory.

% ---- Bibliography ----
%
%kulon bibtex fajl hasznalata
\renewcommand\bibname{\refname}
\renewcommand\bibname{References}
%for papers written in English
\bibliographystyle{splncsnat_en}
\bibliography{main}

\end{document}

### 2023-02-02 mszny2023/transformer-based-huspacy-preprocessing-pipelines/main.tex

\documentclass{llncs}
% \usepackage[round,authoryear]{natbib}

%Magyar nyelvi támogatás (Babel 3.7 vagy késõbbi kell!)
\usepackage[a4paper,includeheadfoot,top=1.65in,bottom=1.65in,left=1.73in,hcentering]{geometry}
\usepackage[pdftex]{graphicx}
\usepackage[sectionbib]{natbib}

\usepackage[utf8]{inputenc}
\usepackage{t1enc}
\usepackage{tikz-dependency}
\usepackage{qtree}
\usepackage{float}
\restylefloat{figure}
\usepackage{siunitx}
\usepackage{tikz-qtree}
\usepackage{todonotes}
% \usepackage[magyar]{babel}
% \selectlanguage{magyar}
\usepackage[hungarian]{babel}
\selectlanguage{hungarian}

% \usepackage[table]{xcolor}

% A formai kovetelmenyekben megkövetelt Times betûtípus hasznalata:
% \usepackage{times}

%A fejléc láblécek kialakításához:
% \usepackage{fancyhdr}

%képekhez
% \usepackage{setspace}
% \usepackage{indentfirst}
\usepackage{graphicx}
\graphicspath{{./figures/}}
% \usepackage{wrapfig}

% ?
% \usepackage{xcolor}
% \usepackage[unicode]{hyperref}
% \usepackage[all]{hypcap}

\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{
            colorlinks,
            linkcolor=black,
            citecolor=blue,
            filecolor=black,
            urlcolor=black,
            pdftitle={Transformer-alapú HuSpaCy előelemző láncok},
            pdfsubject={Transformer modellek magyar nyelvre},
            pdfauthor={},
            pdfkeywords={nlp, spacy, huspacy, transformer, bert, magyar nyelv}}

\usepackage{soul}

\newcommand{\embert}{\texttt{emBERT}}
\newcommand{\emtsv}{\texttt{emtsv}}
\newcommand{\hubert}{\texttt{huBERT}}
\newcommand{\huspacyl}{\texttt{HuSpaCy-large}}
\newcommand{\roberta}{\texttt{XLM-RoBERTa}}
\newcommand{\robertaB}{\texttt{XLM-RoBERTa-base}}
\newcommand{\robertaL}{\texttt{XLM-RoBERTa-large}}
\newcommand{\robertahub}{\texttt{XLM-RoBERTa-base-hu}}
\newcommand{\robertahul}{\texttt{XLM-RoBERTa-large-hu}}
\newcommand{\magyarlanc}{\texttt{magyarlanc}}
\newcommand{\bert}{\texttt{BERT}}
\newcommand{\bertmulti}{\texttt{BERT-base-Multilingual}}
\newcommand{\udpipe}{\texttt{UDPipe}}
\newcommand{\stanza}{\texttt{Stanza}}
\newcommand{\udify}{\texttt{UDify}}
\newcommand{\ud}{Universal Dependencies}
\newcommand{\szc}{Szeged Korpusz}
\newcommand{\spacy}{spaCy}
\newcommand{\huspacy}{HuSpaCy}
\newcommand{\trf}{transformer}

\hyphenation{Stanza}
\hyphenation{emBERT}
\hyphenation{Universal}
\hyphenation{Dependencies}
\hyphenation{emtsv}
\hyphenation{huBERT}
\hyphenation{XLM-RoBERTa}
\hyphenation{XLM-RoBERTa-base}
\hyphenation{XLM-RoBERTa-large}
\hyphenation{XLM-RoBERTa-base-hu}
\hyphenation{XLM-RoBERTa-large-hu}
\hyphenation{magyarlanc}
\hyphenation{spaCy}
\hyphenation{UDPipe}
\hyphenation{UDify}
\hyphenation{Szeged Korpusz}
\hyphenation{Stanza}
\hyphenation{HuSpaCy}
\hyphenation{transformer-alapú}
\hyphenation{BERT}
\hyphenation{BERT-base-Multilingual}

\begin{document}

\pagestyle{myheadings}
\def\leftmark{{\rm XVIII. Magyar Számítógépes Nyelvészeti Konferencia}}
\def\rightmark{{\rm Szeged, 2023. január 27-28.}}

\title{Transformer-alapú \huspacy{} előelemző láncok}

\author{
Szabó Gergő,
Orosz György,
Szántó Zsolt, \break
Berkecz Péter,
Farkas Richárd\\
\institute{
    Szegedi Tudományegyetem, Informatikai Intézet\break
    Szeged, Árpád tér 2.\break
    {\{gszabo,szantozs,berkecz,rfarkas\}@inf.u-szeged.hu}\break
    {gyorgy@orosz.link}\break
}
}

\maketitle

\begin{abstract}
Évről évre jelennek meg egyre pontosabb nyílt forráskódú ter-
mészetes-nyelvfeldolgozó \trf{}-alapú nyelvi modellek. Viszont ezekre épülő, kifejezetten magyar nyelvre fejlesztett előelemző lánc, ami tartalmaz tokenizálást, mondatra bontást, lemmatizálást, szófaji egyértelműsítést, morfológiai címkézést, függőségi elemzést és névelem-felismerést, idáig nem létezett. Dolgozatunkban bemutatunk egy, a fent említett feladatokat megvalósító \trf{}-alapú előelemző láncot, amit több különböző magyarra elérhető nyelvi modellel is teszteltünk. Az elkészült rendszer a feladatok nagy részében state-of-the-art eredményeket ért el. Arra is kellő figyelmet fordítottunk, hogy megvizsgáljunk különböző erőforrás-igényű lehetőségeket és bemutassunk olyan módszereket, amelyekkel a leghatékonyabbnak bizonyult rendszerek memóriahasználata csökkenthető. 
\\{\bf Kulcsszavak:} transformer, huspacy, nyelvi előfeldolgozás
\end{abstract}

\section{Bevezetés}

Az elmúlt években nagyon elterjedtek a \trf{}-alapú modellek alkalmazása (például: \bert{} \citep{devlin2018bert}). Ezek a mély neuronhálókat használó, nagy mennyiségű szövegen előtanított nyelvi modellek azonnal state-of-the-art eredményeket értek el a természetes-nyelvfeldolgozás különböző feladatain \citep{devlin2018bert}. Bár már magyar nyelvre is elérhetők ilyen modellek \citep{Nemeskey:2021a} \citep{conneau2019unsupervised}, olyan ezekre épülő, kifejezetten a magyar nyelv igényeit figyelembe vevő általános előelemző keretrendszer eddig nem készült, ami egy architektúraként foglalná össze a tokenizálást, a mondatra bontást, a szófaji egyértelműsítést, a morfológiai elemzést, a lemmatizálást, a függőségi elemzést és a névelem-felismerést. Célunk egy ilyen eszköz létrehozása volt.

Elemzőnk alapjául a \spacy{} \citep{spacy} keretrendszer magyar nyelvű változatát, a \huspacy{}-t \citep{huspacy} választottuk. A \huspacy{} egy Python nyelvre építő, nyílt forráskódú, szabadon felhasználható nyelvfeldolgozási keretrendszer, ami a korábban felsorolt komponensek mindegyikét tartalmazza. Cikkünkben bemutatjuk, hogy a \huspacy{}-ben található szóbeágyazások lefedő konvolúciós rétegekre épülő neuronháló esetében, milyen javulás érhető el az előtanított \trf{}-alapú modellek beépítésével. Kísérleteinkhez a kizárólag magyar adaton tanított \hubert{}-et \citep{Nemeskey:2021a} és két többnyelvű modellt, a Multilingual BERT-et \citep{pires2019multilingual} és az \roberta{}-t \citep{conneau2019unsupervised} használtuk.

A \trf{} modellek hátránya a nagy számításikapacitás- és a memóriaigényük. Utóbbi kifejezetten igaz a többnyelvű modellek esetén, hiszen ezen modellek szótára óriási mennyiségben tartalmaz olyan szótöredékeket, amik a magyar nyelvben nem fordulhatnak elő. Ennek a problémának a kezelésére megszűrtük az \roberta{} szótárát és megvizsgáltuk, hogy ez milyen hatással van a rendszer teljesítményére.

A további hatékonyság javítás érdekében a \huspacy{} címkézésért felelős neurális architektúrájában két ponton változtattunk. A korábban használt átmenet-alapú függőségi elemzőt kicseréltük egy gráf-alapú változatra, ezen felül a névelem-felismerőt is bővítettük a beam search \citep{beam} beépítésével.

\section{Kapcsolódó munkák}

Kutatásunk fő célja, hogy megvizsgáljuk, hogy a legújabb \trf{}-alapú megoldások hogyan teljesítenek a korábbi megoldásokkal szemben. A magyar nyelvre már korábban is fejlesztettek teljes előelemző láncokat. Ilyen például a \huspacyl{} \citep{huspacy} modellje, amely a WebCorpus 2.0-án \citep{Nemeskey:2020} tanított statikus szóbeágyazásokat használ, illetve egy 4 rétegű CNN-t (\emph{Convolutional Neural Networks}); a \magyarlanc{} \citep{magyarlanc}, amely a rejtett Markov modell-alapú PurePos-t \citep{orosz2013purepos} használja és egy gráf-alapú függőségi címkézőt; a \stanza{} \citep{qi2020stanza}, amely LSTM-alapú (\emph{Long Short Term Memory}) és amely magas pontszámokat ért el az UD korpuszok terén; az \udpipe{} \citep{udpipe}, amelyet alaprendszerként használnak a CoNLL versenyeken; a \udify{} \citep{udify}, ami egy többnyelvű \bert{}-alapú modell és az \emtsv{} \citep{emtsv1, emtsv2, emtsv3, emtsv4, emtsv5}, amely kifejezetten a magyar nyelv igényeire lett fejlesztve, illetve amibe bele lett integrálva az \embert{} \citep{nemeskey2020embert} \trf{}-alapú NER rendszere. A feladatok tekintetében a \huspacy{} és a \stanza{} képes elemezni az összes eddig taglalt feladatot. Ezen túl az \emtsv{} is szintén, viszont a függőségi elemzője nem \ud{} formátumú. Az \udpipe{}-ból és a \magyarlanc{}-ból a névelem-felismerő maradt ki.

\section{Rendszer felépítése}

\subsection{Alapmodell felépítése}
\label{transfer_learn}

Rendszerünk alapja a \huspacy{} \citep{huspacy} cikkben bemutatott, többfeladatos tanulásra építő architektúra képezi, ami egy lépésben kezeli a különböző címkézési feladatokat. Az ebben szereplő 4 rétegű konvolúciós hálókra építő architektúrát lecseréltük egy előtanított \trf{}-alapú modellre.

\begin{figure}[h]
    \center
    \includegraphics[scale=.40]{Transfer_learning.drawio.pdf}
    \caption{Transfer learning ábrázolása.}
    \label{transfer_learning}
\end{figure}

Ahogy az \ref{transfer_learning}. ábrán is látható, a rendszerünk 3 különálló modellből áll. Az első, Szeged Korpuszon \citep{szc} tanított modell végzi a szófaji egyértelműsítést, lemmatizálást és a morfológiai címkézést. Ezt a modellt lemásoltuk két példányban és továbbtanítottuk az egyiket függőségi elemzésre kiegészítve a mondatra bontás feladattal is, a másikat pedig névelem-felismerésre. Erre azért volt szükség, mert az egyes feladatok tanítására eltérő adatbázisok álltak rendelkezésünkre és így mégis képesek voltunk kihasználni az egyes részfeladatok közötti kapcsolatokat. Függőségi elemzésre a Szeged Korpusznál kisebb, viszont nemzetközileg elterjedt címkekészletet használó magyar Universal Dependencies \citep{nivre2017universal} (UD) korpuszt használtuk. A névelem-felismerőt pedig a NerKor \citep{nerkor} és a SzegedNER \citep{szegedner} összevont korpuszán tanítottuk.

Ez a módszer abban tér el a korábbi \huspacy{} architektúrától, hogy csak 2 modell készült. A Szeged Korpuszon tanult modell tovább volt tanítva az Universal Dependencies korpuszon, amely így már ki volt egészítve a függőségi elemző fejjel. Így nem volt egy külön modell a függőségi elemzésre.
A változtatásra a \trf{} architektúra esetén azért volt szükség, mert a függőségi elemző tanítása közben azt tapasztaltuk, hogy a modell elfelejtette a korábban tanult feladatokat. A lemmatizálás, szófaji egyértelműsítés és morfológiai címkézés feladatok esetén a pontosság lezuhant az első néhány epochban, utána javult minimálisan, de például a morfológiai címkézés esetén 6, a lemmatizálás esetén pedig 27 százalékponttal romlott a pontosság a Szeged Korpuszon tanult modellhez képest.

% wandb biaffine_with_all_transfer keyword: lemma_acc|sents_f|pos_acc|morph_acc

\subsection{Az alapmodell kiegészítései}
\label{biaffine}

A \huspacy{} rendszer által alapértelmezettként használt átmenet-alapú függőségi elemzőt lecseréltük a \spacy{} rendszer által biztosított gráf-alapú, biaffine figyelmi \emph{(attention)} mechanizmussal rendelkező elemzőre \citep{biaffine}. Az eredeti cikkben alkalmazott Bi-LSTM rétegek helyett, ahogy a korábbi feladatokban, itt is \trf{}-alapú architektúrát használtunk.
Az átmenet-alapú rendszerek a függőségi fát osztályozási problémák sorozataként építik fel, amikben lehetőség van két vizsgált szó között él behúzására vagy balról jobbra haladva a vizsgált szavak léptetésére. Ezek a megoldások a mohó, lokális döntéseik miatt rosszul tudnak teljesíteni hosszú mondatok esetén. Ezzel szemben a gráf-alapú elemzők a szavakból, mint csúcsokból felírt teljes gráfban keresik a maximális feszítőfát.

A névelem-felismerő a függőségi elemzőhöz hasonlóan egy átmenet-alapú megoldást alkalmaz, amit beam search használatával bővítettünk. Ennek lényege, hogy az elemző az egyes lépésekben több korábbi elemzést is figyelembe vesz. A lemmatizálás feladatban a \cite{peti} cikkben használt megoldásokat alkalmaztuk.

A kifejezetten magyar adatokon tanított \hubert{} mellett kísérleteztünk más, többnyelvű modellekkel is, amelyek támogatják a magyar nyelvet is. Az egyik a \bertmulti{} volt, illetve az \roberta{} két különböző méretű változata, az \robertaB{} és az \robertaL{}.

A \hubert{}, a \bertmulti{} és az \robertaB{} mindegyike 12 rétegből áll, amelyek egyenként 768 dimenziós vektorokat használnak. Az előbbi kettő összesen 110 millió paraméterrel rendelkezik, az \robertaB{} pedig 125 millióval. Ezzel szemben az \robertaL{} kétszer annyi réteget és 1024 dimenziós vektorokat használ és összesen 355 millió paraméterrel rendelkezik.

\subsection{Többnyelvű modellek méretcsökkentése}
\label{roberta_size}

A többnyelvű modellek egyik hátránya a nagyobb memóriaigényük, aminek az oka, hogy az eltérő nyelvek szavai miatt sokkal nagyobb szótárra van szükségük. Egynyelvű felhasználás esetén viszont a szótár egy jelentős részére szinte biztosan nincs szükségünk. Ilyenek például a kínai szimbólumok, török betűk, cirill betűk és az azokra épülő szavak, szótöredékek. Ennek kezelésére reguláris kifejezésekkel megszűrtük a két \roberta{} modell szótárát. A jelenlegi szűrés csak magyar nyelvben használt betűket és írásjeleket tartalmazó szótöredékeket hagyta meg.

\section{Eredmények}

Az összes kiértékelés (a névelem-felismerő kivételével) a Hungarian Universal Dependencies Corpus
tesztkészletén \citep{UniDep} készült és a CoNLL 2018 Shared Taskon\footnote{\url{https://universaldependencies.org/conll18/conll18_ud_eval.py}} alkalmazott kiértékelő szkript felhasználásával történt. A nem \trf{}-alapú modellek esetén összehasonlításképpen öt rendszert választottunk. Az első az \emtsv{} \citep{emtsv1, emtsv2, emtsv3, emtsv4, emtsv5}, második az \udpipe{} \citep{udpipe}, a harmadik a \stanza{} \citep{qi2020stanza} és a negyedik a \udify{} \citep{udify}. Ugyanakkor a 2021-ben publikált \huspacyl{} \citep{huspacy} modelljével is összehasonlítottuk.

A \stanza{}\footnote{\url{https://stanfordnlp.github.io/stanza/performance.html}} és \udpipe{}\footnote{\url{https://ufal.mff.cuni.cz/udpipe/1/models}} szerzői már kimérték a rendszerüket az UD teszthalmazon, de ez nem mondható el az \emtsv{}-ről. Itt a rendszert az alapértelmezett konfigurációs beállítások megtartásával lefuttattuk az általunk alkalmazott teszthalmazon. Ugyanakkor fontos kiemelni, hogy az \emtsv{} eredményei olyan szempontból nem számítanak összevethetőnek, hogy nem lett újratanítva az általunk használt adatbázis vágásokkal, hanem csak egyszerű kiértékelés történt. Így a kiértékelés során előfordulhatott átfedés a tanító és a kiértékelő adathalmazok között. A \udify{} kiértékelése során gold adatot használt a tokenizálás és a mondatra bontás feladatokban.

A \trf{}-alapú modellek esetében felhasználtuk az eddigi tapasztalatainkat és további kísérleteket végeztünk néhány egyéb előtanított nyelvi modellel a \huspacy{} mostanra már alapértelmezett konfigurációjával. A következőkben részletesen bemutatjuk az eredményeket, valamint azt is, hogy mely nyelvi modellekkel sikerült elérni őket.

\subsection{Tokenizálás és mondatra bontás}

A mondatra bontás esetén látható, hogy minimálisan, de a \trf{} modellek meghaladják a nem \trf{}-alapú modelleket. Továbbá látható, hogy az \robertaL{}-alapú modell szinte tökéletesen szegmentált, így megelőzve társait.
A tokenizálás semmilyen technikában nem tért el a \huspacy{} cikkben ismertetett módszerektől.

\newlength{\lsz}
\settowidth{\lsz}{Mondatra bontás}
\begin{table}[h]
    % \vspace{2em}
    \begin{center}
        \begin{tabular}{
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lsz}
            >{\centering\arraybackslash}m{\lsz}}
            \toprule
                                      & Token               & Mondatra bontás \\
            \midrule
            \stanza{}                 & \textbf{99,92\%}    & 97,45\%         \\
            \udpipe{}                 & 99,80\%             & 95,90\%         \\
            \emtsv{}                  & 99,77\%             & 98,67\%         \\
            \udify{}                  & --                  & --             \\
            \huspacyl{}               & 99,89\%             & 98,55\%         \\
            \midrule
            \hubert{}                 & 99,89\%             & 99,33\%        \\
            \bertmulti{}              & 99,89\%             & 87,75\%        \\
            \robertaB{}               & 99,89\%             & 99,33\%        \\
            \robertaL{}               & 99,89\%             & \textbf{99,67\%}        \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{Mondatra bontás F1-score eredményei az UD teszthalmazán.}
        \label{table:tokens}
    \end{center}
    \vspace{-3em}
\end{table}

\subsection{Morfo-szintaktikai elemzés}

A \ref{table:tagging}. táblázat eredményei alapján elmondható, hogy a \hubert{}-alapú modell kiemelkedően teljesít minden területen. A morfológiai elemzésben sikerült több, mint fél százalékponttal javítani az eddigi legjobb eredményen. A szófaji címkézés esetében szintén sikerült javítani több, mint egy százalékponttal. Ezen felül a lemmatizálás feladatban is legalább 4 százalékponttal javított minden eddigi rendszerhez képest.

Ugyanakkor összesítve az \robertaL{}-alapú modell ismét majdnem minden más modellt (lemmatizálásnál szinte fej fej mellett van a \hubert{}-alapú modellel) megelőz, morfológiai elemzés esetén is 3 százalékponttal felülmúlja a \hubert{}-alapú modellt.

\newlength{\lm}
\settowidth{\lm}{Morph. acc.}
\begin{table}[h]
    % \vspace{2em}
    \begin{center}
        %\addtolength{\tabcolsep}{-5pt}
        \begin{tabular}{
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lm}
            >{\centering\arraybackslash}m{\lm}
            >{\centering\arraybackslash}m{\lm}
            }
            \toprule
                                     & Lemma pontosság       & PoS pontosság    & Morf. pontosság  \\
            \midrule
            \stanza{}                & 94,19\%               & 96,00\%          & 93,62\%    \\
            \udpipe{}                & 88,50\%               & 90,60\%          & 88,50\%            \\
            \emtsv{}                 & 96,16\%               & 89,19\%          & 89,12\%           \\
            \udify{}                 & 90,19\%               & 96,36\%          & 86,16\%          \\
            \huspacyl{}              & 97,46\%               & 96,89\%          & 93,87\%           \\
            \midrule
            \hubert{}                 & \textbf{98,68\%}    & 97,27\%           & 94,20\%         \\
            \bertmulti{}              & 98,48\%             & 97,11\%           & 91,76\%         \\
            \robertaB{}               & 98,55\%             & 97,73\%           & 96,61\%         \\
            \robertaL{}               & 98,67\%             & \textbf{98,01\%}  & \textbf{97,15\%} \\

            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{Lemmatizálás, szófaji címkéző (PoS) és morfológiai elemző F1-score eredményei az UD teszthalmazán.}
        \label{table:tagging}
    \end{center}
    \vspace{-2em}
\end{table}

\subsection{Gráf-alapú függőségi elemző}

A \ref{table:parser}. táblázatban látható az átmenet-alapú függőségi elemző, a gráf-alapú függőségi elemző,
illetve a többi rendszer eredményei. Mivel az \emtsv{} nem szolgáltat \ud{} formátumú függőségi
elemzést, így ez a modell nem mérhető ezzel a feladattal.

\newlength{\lmm}
\settowidth{\lmm}{Parser}
\begin{table}[h]
    % \vspace{-1em}
    \begin{center}
    \setlength{\tabcolsep}{10pt}
        \begin{tabular}{
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lmm}
            >{\centering\arraybackslash}m{\lmm}
            }
            \toprule
                                       & UAS                      & LAS       \\
            \midrule
            \stanza{}                  & 84,19\%                  & 79,23\%   \\
            \udpipe{}                  & 72,80\%                  & 67,20\%   \\
            \udify{}                   & 89,69\%                  & 84,88\%          \\
            \huspacyl{}                & 82,53\%                  & 75,56\%   \\
            \midrule
            \hubert{} - átmenet-alapú  & 89,95\%                  & 83,94\%  \\
            \hubert{} - gráf-alapú     & \textbf{91,57\%}         & \textbf{87,58\%}  \\
            \midrule
            \bertmulti{} - gráf-alapú              & 85,92\%                 & 81,93\%           \\
            \robertaB{} - gráf-alapú               & 89,47\%                 & 86,02\%           \\
            \robertaL{} - gráf-alapú               & 90,38\%                 & 86,88\%          \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{Átmenet-alapú és gráf-alapú függőségi elemzés összehasonlítása a meglévő rendszerekkel, illetve a magyar nyelvre elérhető \trf{}-alapú nyelvi modellek kiértékelése.}
        \label{table:parser}
    \end{center}
    \vspace{-3em}
\end{table}

A meglévő rendszerek és a \huspacyl{} modellje is alulmarad a \hubert{} átmenet-alapú modellel szemben, mind az UAS és a LAS tekintetében is. Az átmenet-alapú függőségi elemzővel rendelkező \hubert{}-alapú modell viszont alulmarad a már tárgyalt gráf-alapú megoldással szemben. Ezért ezt alkalmaztuk is a továbbiakban, illetve a többi \trf{}-alapú modellnél is ezt használtuk alapértelmezett elemzőként függőségi elemző esetén.

Viszont ebben a feladatban az \robertaL{} pontatlanabb a \hubert{} gráf-alapú modellel szemben. Ez az egyetlen olyan komponens, ahol a \hubert{}-alapú modell sokkal jobban teljesít az \robertaL{}-alapú modellnél.

\subsection{Névelem-felismerés}

Mivel az UD korpusz nem tartalmaz névelemeket, ezért erre a célra a NerKor \citep{nerkor} és a SzegedNER \citep{szegedner} korpuszt használtuk. Mivel az \udpipe{}-ban nincs névelem-felismerő, így azt nem lehetett bevonni a vizsgálatba. Három korábbi névelem-felismerőt is bevettünk az összehasonlításba. Az egyik a \cite{szarvas-ner}, amely döntési fákat használ. A második a \cite{simon-ner}, amely egy lineáris modellt használ és rejtett Markov modelleket kombináló statisztikai címkézővel rendelkezik. A harmadik, újabb rendszer pedig az \embert{}, amely egy \trf{}-alapú NER modell. A \stanza{} esetén viszont fontos kiemelni, hogy újra lett tanítva az általunk használt adatbázisokkal \citep{nerkoreval}.

\newlength{\lnersz}
\settowidth{\lnersz}{SzegedNer}
\begin{table}[h]
    \begin{center}
        \begin{tabular}{
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lnersz}
            >{\centering\arraybackslash}m{\lnersz}
            >{\centering\arraybackslash}m{\lnersz}
            }
            \toprule
                                    & SzegedNER & NerKor  & Kombinált        \\
            \midrule
            \cite{simon-ner}        & 95,06\%   & --      & --               \\
            \cite{szarvas-ner}      & 94,77\%\  & --      & --               \\
            \embert{}               & 97,40\%   & 92,09\% & \textbf{92,99\%} \\
            \stanza{}               & 91,78\%   & 80,53\% & 83,75\%          \\
            \huspacyl{}             & 95,31\%   & 80,75\% & 85,73\%          \\
            \midrule
            \hubert{}                     & 97,01\%    & 88,27\% & 90,26\%   \\
            \hubert{} + beam search       & 97,37\%    & 89,13\% & 91,27\%   \\
            \midrule
            \bertmulti{} + beam search              & --    & --   & 89,07\%         \\
            \robertaB{} + beam search               & --    & --   & 90,94\%         \\
            \robertaL{} + beam search               & --    & --   & \underline{91,86\%}   \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{F1-scoreon mért névelem-felismerés összehasonlítása a SzegedNER, a NerKor és a kombinált teszthalmazon is.}
        \label{table:ner_cmp}
    \end{center}
    \hspace*{.1\linewidth}
    \vspace{-4em}
\end{table}

A NerKor adatbázis esetében a szerzők által meghatározott tanító-validáló-teszt halmazokra
bontást alkalmaztuk. A SzegedNER esetében pedig, az összehasonlítás érdekében, a \cite{szarvas-ner} cikk által meghatározott vágás lett felhasználva.

A meglévő rendszerek, az \embert{} kivételével, alulmaradnak a \huspacyl{} modellel és a \hubert{}-alapú modellel szemben. Viszont a \hubert{}-alapú + beam search modell jobb eredményt ért el, mint ahol nem volt alkalmazva a beam search, ahogy a \ref{table:ner_cmp}. táblázatban is látható. A konklúzió tehát, hogy jobb a beam search, ezért a többi \trf{}-alapú modellnél is ezt használtuk alapértelmezett módszerként.

Így, mivel bevettük a vizsgálatba a régebbi rendszereket, ezért külön korpuszokra is le lettek mérve a modellek. Az \embert{}-tel való összehasonlítás során a SzegedNER korpusz esetében szinte fej fej mellett van a két modell (\embert{} és a beam search \hubert{}-alapú modell), de a NerKor esetében kicsit alulmarad a \hubert{}-alapú modell. A kombinált korpuszon is jobb az \embert{}, mert a 12 réteg feletti Viterbi-algoritmus garantálja a jósolt címkeszekvencia konzisztenciáját. Ezen felül, ha megnézzük az összes \trf{}-alapú modellt és az \embert{} eredményeit, kijelenthető, hogy az utóbbi továbbra is a leghatékonyabb.

\subsection{\roberta{}-hu modell méretcsökkentése}

Ahogyan a \ref{roberta_size}. fejezetben említettük, az \roberta{} modellek méretét, ha sikerülne csökkenteni megtartva a teljesítményét, akkor elérhető lenne egy rendkívül pontos modell jelentősen kevesebb memóriaigénnyel.

Az első fázisban reguláris kifejezések segítségével szűrtük ki azokat a tokeneket, amelyek nem fellelhetőek a magyar és az angol nyelvben. Így például a kínai szimbólumokat, török betűket, cirill betűket tartalmazó részszavakat és karaktereket kivettük. Ugyanakkor az Embedding layerből is eltávolítottuk a hozzátartozó vektorokat, így a paraméterszáma is csökkent a modellnek.

\newlength{\lroberta}
\settowidth{\lroberta}{98,56\%aa}
\newlength{\lrobertalem}
\settowidth{\lrobertalem}{Lemmatizálás}
\begin{table}[h]
    % \vspace{-2em}
    \begin{center}
    \setlength{\tabcolsep}{1pt}
        \begin{tabular}
            {
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lroberta}
            >{\centering\arraybackslash}m{\lroberta}
            >{\centering\arraybackslash}m{\lroberta}
            >{\centering\arraybackslash}m{\lroberta}
            >{\centering\arraybackslash}m{\lrobertalem}
            }
            \toprule
                            & Token                              & Mondatra bontás                & PoS                 & Morf.               & Lemma          \\
            \midrule
            \robertaB{}     & \textbf{\multirow{2}{*}{99,89\% }} & \underline{99,33\%}  & \underline{97,73\%} & \underline{96,61\%} & \underline{98,55\%} \\
            \robertahub{}   &                                    & 98,56\%              & 97,60\%             & 96,27\%             & 98,42\%             \\
            \midrule
            \robertaL{}     & \textbf{\multirow{2}{*}{99,89\% }} & \textbf{99,67\%}     & \textbf{98,01\%}    & \textbf{97,15\%}    & \textbf{98,67\%}    \\
            \robertahul{}   &                                    & 97,24\%              & 97,46\%             & 95,84\%             & 98,11\%             \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{Módosított \roberta{} modellek F1-score eredményei az UD korpuszon. (Az aláhúzás az aktuális résztáblázatban a legjobb eredmény, a félkövérített pedig az adott egész táblázatban a legjobb eredmény.)}
        \label{table:roberta}
    \end{center}
    % \hspace*{-4\linewidth}
    \vspace{-3em}
\end{table}

\newlength{\lrobertaa}
\settowidth{\lrobertaa}{XLMRoBERTa2}
\begin{table}[h]
    % \vspace{-2em}
    \begin{center}
    % \setlength{\tabcolsep}{10pt}
        \begin{tabular}
            {
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lrobertaa}
            >{\centering\arraybackslash}m{\lrobertaa}
            >{\centering\arraybackslash}m{\lrobertaa}
            }
            \toprule
                            & UAS                 & LAS                 & NER        \\
            \midrule
            \robertaB{}     & \underline{89,47\%} & \underline{86,02\%} & \underline{90,94\%}  \\
            \robertahub{}   & 89,21\%             & 85,54\%             & 89,89\%           \\
            \midrule
            \robertaL{}     & \textbf{90,38\%}    & \textbf{86,88\%}    & \textbf{91,86\%}  \\
            \robertahul{}   & 89,40\%             & 85,46\%             & 90,44\%           \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{Módosított \roberta{} modellek UAS és LAS eredményei az UD korpuszon, a NER pedig a kombinált NER adatbázison lett mérve, amely F1-scoret mutatnak.}
        \label{table:roberta2}
    \end{center}
    % \hspace*{-4\linewidth}
    \vspace{-3em}
\end{table}

Ahogy a \ref{table:roberta}. és a \ref{table:roberta2}. táblázatokban is észrevehető, a méretoptimalizált modellekkel hozzávetőlegesen sikerült tartani a teljesítményt. Jelenlegi kísérletek is hasznos eredményeknek számítanak, ugyanis az \robertahub{} esetén például sikerült a \textbf{felére csökkenteni} a méretet és az \robertahul{} modell pedig megközelítőleg \textbf{kétharmada az eredetinek}. Ezt a következő fejezetben jobban kifejtjük és összehasonlítjuk. Konklúzió, hogy a jelenlegi módszer segítségével képesek vagyunk csökkenteni a \roberta{} modell méretét minimális pontosságvesztéssel. Annak érdekében, hogy valóban hasznos legyen ez a modell és megtartsa a pontosságot, szükség lenne egy olyan heurisztikára, amely képes lenne ezt a feladatot ellátni. Jövőbeli tervünk, hogy az eldobandó szótöredékek halmazát statisztikai alapon válasszuk ki. Ehhez szeretnénk felhasználni a WebCorpus 2.0-át \citep{Nemeskey:2020} és csak az abban előforduló leggyakoribb szótöredékeket megtartani, amivel a terveink szerint tovább csökkenthetnénk a szótár méretét és megtarthatnánk olyan speciális karaktereket is, amiket a kézi szabályokkal nem fedtünk le. A jelenlegi script elérhető publikus módon a GitHubon\footnote{\url{https://github.com/huspacy/huspacy-resources/blob/master/scripts/XLM-RoBERTa_size_reduction/XLM-RoBERTa_size_reduction.ipynb}}, ami lehetővé teszi egy új specifikus nyelvi modell készítését bármely nyelvre.

\subsection{Memóriahasználat}

A \trf{} architektúrák esetén fontosak az erőforrásigények, ezért ebben a fejezetben mélyebben megvizsgáljuk a futásidőket és a memóriaigényeket az összes eddigi modell szemszögéből. Az \udpipe{}, \emtsv{} és az \udify{} rendszerek nem támogatják a névelem-felismerést, így ezeket a modelleket a névelem-felismerő nélkül mértük. A mérések az UD teszthalmazán készültek.

\newlength{\lper}
\settowidth{\lper}{Throughput tok}
\begin{table}[h]
    \begin{center}
    % \setlength{\tabcolsep}{1pt}
        \begin{tabular}{
            l<{\hspace{1em}}
            >{\centering\arraybackslash}m{\lper}
            >{\centering\arraybackslash}m{\lper}
            >{\centering\arraybackslash}m{\lper}
            }
            \toprule
                                      & Gyorsaság (token/mp) CPU & Gyorsaság (token/mp) GPU & Memóriahasználat (GB) \\
            \midrule
            \stanza{}                 & 30         & 395        & 5,3  \\
            \udpipe{}                 & 3175       & --         & 1,3  \\
            \emtsv{}                  & 113        & --         & 3,9  \\
            \udify{}                  & 129        & 475        & 3,2  \\
            \huspacyl{}               & 728        & 4685       & 3,2  \\
            \midrule
            \hubert{}                 & 176        & 2605        & 4,8  \\
            \bertmulti{}              & 138        & 2631        & 7,0  \\
            \robertaB{}               & 151        & 2847        & 11,2 \\
            \robertahub{}             & 166        & 3265        & 6,2  \\
            \robertaL{}               & 50         & 2353        & 18,9 \\
            \robertahul{}             & 59         & 2390        & 14,6 \\
            \bottomrule
        \end{tabular}
        \vspace{1em}
        \caption{A CPU-n (AMD EPYC 7F72) és GPU-n (NVIDIA A100 40 GB) összehasonlított rendszerek gyorsasága (token/másodpercben mérve) és a maximális memóriahasználatuk.}
        \label{table:performance}
    \end{center}
    \vspace{-3em}
\end{table}

Először nézzünk meg a CPU-s oszlopot. A \ref{table:performance}. táblázatban jól látható, hogy a \trf{} modellek a többi rendszerekhez képest több erőforrást igényelnek. Viszont kivételt képez például az \emtsv{} gyorsasága a \hubert{}-alapú modellhez képest. A \stanza{} szintén lassabb, memóriaigények tekintetében pedig egy skálán mozognak. Az \udpipe{}
ellenben mind a két esetben jól teljesített a futásidő és memóriaigények tekintetében, ugyanakkor ez a rendszer minden elemző esetén alulmarad hatékonyságban a többi rendszerhez képest. A többi \trf{} modell több paraméterrel rendelkezik, ezért és a többnyelvűség miatt is több erőforrást igényelnek.

Megfigyelhető az is, hogy a \hubert{}-alapú modell memóriahasználatban a \huspacyl{} modelljéhez képest nem mutat jelentős eltérést. Ellenben az \roberta{} modellekre ez már nem teljesen igaz. Ahogyan szó volt a \ref{transfer_learn}. fejezetben a modellek összetételéről, itt nyilvánul meg a három \trf{} modell mérete és az, hogy a többnyelvű modellek sokkal több memóriát igényelnek már önmagukban is.

A méretoptimalizált és az eredeti modellek között a futásidő azért nem változott drasztikusan, mert tulajdonképpen mind a két esetben ugyanazok a neuronok feleltek a futásidőért. Más szóval, a modell csak azon neuronokat használta fel a predikáláshoz, amelyekre a magyar nyelvhez szükség van. Az, hogy el lettek távolítva azok a részek, amelyeket a modell egyébként sem használt, nem befolyásolta a futásidőt.

\section{Összegzés}

Elkészítettünk a \huspacy{}-hez\footnote{\url{https://github.com/huspacy/huspacy/tree/develop}} egy \hubert{}-alapú %\trf{} 
modellt\footnote{\url{https://huggingface.co/huspacy/hu_core_news_trf}}, ami minden
hagyományos nyelvi elemzőt tartalmaz, vagyis tokenizálót, mondatra bontót, lemmatizálót, szófaji és morfológiai egyértelműsítőt, függőségi elemzőt, illetve ezeken felül névelem-felismerőt is és ezen eredmények teljes mértékben reprodukálhatóak.

Összességében a \trf{}-alapú modellek minden feladatban jobb eredményt érnek el a korábbi megoldásoknál. Ezek közül is a legtöbb feladatban a \robertaL{} teljesített a legjobban, csupán a függőségi elemzés részfeladaton múlta felül a \hubert{}-alapú modell. Az alacsonyabb memóriaigényének és gyorsabb futásának köszönhetően a \hubert{} sok esetben jó választás lehet, de, ha a pontosságon van a hangsúly, a \robertaL{} felhasználásával érhetjük el a legjobb eredményt.
Viszont a state-of-the-art eredmények ellenére is hátrányt jelenthet bizonyos végalkalmazásokban a nagyobb erőforrásigényük, például időérzékeny alkalmazásokban, illetve, ha korlátozott erőforrásokkal rendelkezünk. Ezekben a szituációkban megfelelő döntés lehet a \huspacyl{} modell használata.

A függőségi elemzés és a névelem-felismerés problémákra több algoritmust is kipróbáltunk és a legmegfelelőbbet alkalmaztuk. Az eredmények alapján, a \hubert{}-alapú modell a függőségi elemző esetén hatalmas javulást mutat az elődeihez képest, így sikerült egy state-of-the-art eredményt elérni. Továbbá a morfológiai elemzés esetén is az \robertaL{}-alapú modell képest volt 4 százalékponttal javulni az eddigi legjobb eredményhez képest.

Az elkészült modellek bárki számára szabadon igénybe vehetők a nyílt forráskódú \huspacy{} részeként.

\section*{Köszönetnyilvánítás}

A kutatás az Európai Unió támogatásával valósult meg, az
RRF-2.3.1-21-2022-00004 azonosítójú, Mesterséges Intelligencia Nemzeti Laboratórium projekt keretében. Továbbá szeretnénk megköszönni Berend Gábor segítségét a nyelvi modellek méretcsökkentésével kapcsolatban.

\renewcommand\bibname{Hivatkoz\'asok}

\bibliographystyle{splncsnat}
\bibliography{main}

\end{document}

### 2023-06-05 tsd2023/main.tex
feat: added tsd2023 article

%
%	TSD 2023
%	LaTeX Template for Camera-ready Version
%
%	Rel. 2013-05-20 by Ivan Habernal (habernal@kiv.zcu.cz)
%	Rel. 2014-11-21 by Kamil Ekstein (kekstein@kiv.zcu.cz)
%   Rel. 2015-02-23 by Pavel Kral    (pkral@kiv.zcu.cz)
%	Rel. 2017-02-06 by Kamil Ekstein (kekstein@kiv.zcu.cz)
%	Rel. 2018-10-26 by Kamil Ekstein (kekstein@kiv.zcu.cz)
%	Rel. 2021-02-15 by Kamil Ekstein (kekstein@kiv.zcu.cz)
%	Rel. 2023-02-10 by Kamil Ekstein (kekstein@kiv.zcu.cz)
%
%	Based upon Springer's LNCS series template.
%
\documentclass[runningheads,a4paper]{llncs}

\usepackage{times}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{url}

% TSD 2023: Add any additional packages you use in your manuscript
% -----pack
% \usepackage{xxx}
% new packages
\usepackage{array}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{tabularx}
\usepackage{svg}
\usepackage[inline]{enumitem}
\usepackage{todonotes}
\usepackage{float}
\usepackage{hyperref}
\usepackage[british]{babel}
\hypersetup{plainpages=false,
            pageanchor=true,
            colorlinks=true,
            linkcolor=blue,
            citecolor=blue,
            anchorcolor=blue,
            menucolor=blue,
            filecolor=blue,
            urlcolor=blue,
            linktocpage=true,
            breaklinks=true,
            unicode=true}
\usepackage[htt]{hyphenat}

\hyphenation{base-line}
\hyphenation{base-lines}

% -----

% TSD 2023: Add your custom definitions etc., if required
% -----misc
% \newcommand{\xxx}[1]{[#1]}
\newcommand{\spacy}{spaCy}
\newcommand{\huspacy}{HuSpaCy}
\newcommand{\emtsv}{\texttt{emtsv}}
\newcommand{\magyarlanc}{\texttt{magyarlanc}}
\newcommand{\udpipe}{UDPipe}
\newcommand{\stanza}{Stanza}
\newcommand{\trankit}{Trankit}
\newcommand{\udify}{UDify}

\newcommand{\universaldependencies}{Universal Dependencies}
\newcommand{\ud}{UD}
\newcommand{\udhu}{UD-Hungarian}
\newcommand{\szegedcorpus}{Szeged Corpus}
\newcommand{\nerkor}{NYTK-NerKor}
\newcommand{\szegedner}{Szeged NER}

\newcommand{\floret}{\texttt{floret}}
\newcommand{\fasttext}{\texttt{fastText}}
\newcommand{\wordtovec}{\texttt{word2vec}}
\newcommand{\biaffine}{Biaffine}
\newcommand{\lemming}{Lemming}

\newcommand{\fone}{$F_1$}

\newcommand{\hubert}{\texttt{huBERT}}
\newcommand{\embert}{\texttt{emBERT}}
\newcommand{\bert}{\texttt{BERT}}
\newcommand{\xlmroberta}{\texttt{XLM-RoBERTa-large}}

\newcommand{\huspacylg}{\texttt{lg}}
\newcommand{\huspacymd}{\texttt{md}}
\newcommand{\huspacytrf}{\texttt{trf}}
\newcommand{\huspacytrfxl}{\texttt{trf\_xl}}
% -----

\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

% TSD2023: Put your e-mail addresses here
\urldef{\mailsa}\path|gyorgy@orosz.link|
\urldef{\mailsb}\path|{berkecz, gszabo, szantozs, rfarkas}@inf.u-szeged.hu|
%\urldef{\mailsc}\path|other_mails_if_needed|

\begin{document}

% TSD 2023: Put your title here (please, use capitalization, see e.g.
% http://en.wikibooks.org/wiki/Basic_Book_Design/Capitalizing_Words_in_Titles)
\title{Advancing Hungarian Text Processing with \huspacy: Efficient and Accurate NLP Pipelines}

% TSD 2023: a short form should be given in case the title is too long for the running head
\titlerunning{Advancing Hungarian Text Processing with \huspacy}

% TSD 2023: Author's names. Chinese authors should write their first names(s)
% in front of their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
% If the names contain accented characters, please use escape codes
% (refer to http://en.wikibooks.org/wiki/LaTeX/Special_Characters#Escaped_codes)

\author{Gy\"{o}rgy Orosz$^*$ \and Gerg\H{o} Szab\'{o}$^*$ \and P\'{e}ter Berkecz$^*$ \and \\ Zsolt Sz\'{a}nt\'{o} \and Rich\'{a}rd Farkas}

% TSD 2023: For authors from different institutions, please use the following
% form including institution reference
% \author{Firstname1 Surname1\inst{1} \and Firstname2 Surname2 \inst{2}}

% TSD 2023: Author's names for headings. For 1-2 authors, use the following form
% \authorrunning{Firstname1 Surname1 and Firstname2 Surname2}
% TSD 2023: For more than 2 authors, please, use the following

\authorrunning{Orosz et al.}

% TSD 2023: The authors' affiliations

\institute{Institute of Informatics, University of Szeged\break
    2. Árpád tér, Szeged, Hungary \\
% TSD 2023: optional url
% % \url{www.website.org} \\
% \mailsa\\
% \mailsb
% % TSD 2023: For authors from different institutions, add 2nd institution, etc.
% % \and
% % Affiliation2, Institute2, Address \\
% % \url{www.website.org} \\
% % \mailsb\\
\email{gyorgy@orosz.link}\break
\email{\{gszabo,berkecz,szantozs,rfarkas\}@inf.u-szeged.hu}\break
}

% TSD 2023: Put all authors' names to the proceeding index (surname, first name)
\index{Orosz, Gy\"{o}rgy}
\index{Szab\'{o}, Gerg\H{o}}
\index{Berkecz, P\'{e}ter}
\index{Sz\'{a}nt\'{o}, Zsolt}
\index{Farkas, Rich\'{a}rd}

\toctitle{} \tocauthor{}

\maketitle

\def\thefootnote{*}\footnotetext{These authors contributed equally to this work.}\def\thefootnote{\arabic{footnote}}

%
%
%	TSD 2023 SUBMISSION TEXT
%
%
\begin{abstract}
% TSD 2023:
This paper presents a set of industrial-grade text processing models for Hungarian that achieve near state-of-the-art performance while balancing resource efficiency and accuracy. Models have been implemented in the \spacy{} framework, extending the \huspacy{} toolkit with several improvements to its architecture. Compared to existing NLP tools for Hungarian, all of our pipelines feature all basic text processing steps including tokenization, sentence-boundary detection, part-of-speech tagging, morphological feature tagging, lemmatization, dependency parsing and named entity recognition with high accuracy and throughput. We thoroughly evaluated the proposed enhancements, compared the pipelines with state-of-the-art tools and demonstrated the competitive performance of the new models in all text preprocessing steps. All experiments are reproducible and the pipelines are freely available under a permissive license.

% TSD 2023: keywords, comma-separated
\keywords{Hungarian NLP, \spacy, PoS tagging, lemmatization, dependency parsing, named entity recognition}
\end{abstract}

% \todo{Rev1: Some results are presented in a rather sketchy way, abbreviations in the table headings are not always explained. It is also not clear what units were considered in the throughput comparisons.}
% \todo{Rev2: The exact descriptions of the four models' features are mostly scattered around the text - a tabular overview is needed here.
% }

\section{Introduction} \label{sec:introduction}

Academic research in natural language processing has been dominated by end-to-end approaches utilizing pre-trained large neural language models which are fine-tuned for the particular applications. Although these deep learning solutions are highly accurate, there is an important demand for human-readable output in real-world language processing systems. Industrial applications are frequently fully or partially rule-based solutions, as (sufficient) training data for a pure machine learning solution is not available and each and every real-world application has its own requirements. Moreover, rule-based components provide tight control over the behavior of the systems in contrast to other approaches.

%Hand-crafted rules are formed on the top of linguistic features (e.g. PoS tags, named entities) of texts. 
In this paper, we present improvements to a Hungarian text preprocessing toolkit that achieve competitive accuracies compared to the state-of-the-art results in each text processing step. An important industrial concern about large language models is the computational cost, which is usually not worth the accuracy gain. Transformer-based language models require far more computational resources than static word vectors, and their running costs are typically orders of magnitude higher. Furthermore, practical NLP solutions using large language models often only outperform more lightweight systems by a small margin.

In this work, we focus on text processing pipelines that are controllable, resource-efficient and accurate. We train new word embeddings for cost-effective text processing applications and we provide four different sized pipelines, including transformer-based language models, which enable a trade-off between the running costs and accuracy for practical applications. To make our pipelines easily controllable, we implement them in the \spacy\footnote{\url{https://spacy.io/}} framework \cite{spacy} by extending \huspacy{} \cite{huspacy:2021} with new models.

\section{Background} \label{sec:background}
\subsection{Specification for Language Processing Pipelines for Industrial Use} \label{sec:background:specification}

Text processing tools providing representation for hand-crafted rule construction should consist of tokenization, sentence splitting, PoS tagging, lemmatization, dependency parsing, named entity recognition and word embedding representation. These solutions have to be accurate enough for real-world scenarios while they should be resource-efficient at the same time. Last but not least, modern NLP applications are usually multilingual and should quickly transfer to a new language. This can be provided by relying on international annotation standards and by the integration into multilingual toolkits.

\subsection{Annotated Datasets for Preprocessing Hungarian Texts} \label{sec:background:datasets}

According to Simon et al. \cite{metanet}, Hungarian is considered to be one of the best supported languages for natural language processing. In 2004, the \szegedcorpus{} \cite{szc} was created, comprising 1.2 million manually annotated words for part-of-speech tags, morphological descriptions, and lemmata. Subsequently, these annotations were extended \cite{szeged-treebank} with dependency syntax annotations. In 2017, a small section of the corpus was manually transcribed to be a part of the \universaldependencies{} (\ud) project \cite{nivre-etal-2020-universal}. Around the same time, the entire corpus was automatically converted from the original codeset to the universal part-of-speech and morphological descriptions \cite{vincze-etal-2017-universal}.

\szegedner{} \cite{szarvas-ner}, developed in 2006, was the first Hungarian named entity recognition corpus, consisting of 200,000 words of business and criminal news. In recent years, \nerkor{} \cite{nerkor} extended the possibilities of training and benchmarking entity recognition systems for Hungarian with a one million word multi-domain corpus.

\subsection{Multilingual NLP Toolkits} \label{sec:background:toolkits}

Thanks to the \ud{} project, it is now possible to easily construct multilingual NLP pipelines. Among the most commonly utilized toolkits are \udpipe{} \cite{udpipe}, \stanza{} \cite{stanza}, \udify{} \cite{udify}, \trankit{} \cite{van2021trankit} and \spacy.

On the one hand, these systems exhibit a high degree of algorithmic diversity. They can be classified into two distinct groups based on their utilization of neural networks. \udpipe, \spacy{} and \stanza{} apply older, but faster architectures built on word embeddings employing convolutional and recurrent layers, respectively. On the contrary, \udify{} and \trankit{} leverage transformer-based large language models, with the former using multilingual \bert{} \cite{devlin2018bert} while the latter utilizing \xlmroberta{} \cite{conneau2019unsupervised}.

On the other hand, these frameworks are typically limited by the fact that they rely solely on the \universaldependencies{} datasets, which may present a disadvantage in languages such as Hungarian, which have large corpora incompatible with \ud. Each of the above-mentioned systems shares this limitation, moreover, \spacy{} does not offer a Hungarian model at all, due to the restrictive license of the \udhu{} corpus. Regarding named entity annotations, \stanza{} is the only tool supporting NER for Hungarian.

\subsection{Hungarian Language Processing Tools}  \label{sec:background:hunlp}

The landscape of the Hungarian text processing systems was similar to that of English before the ``industrial NLP revolution''. There were a number of standalone text analysis tools \cite{metanet} capable of performing individual text processing tasks, but they often did not work well with each other.

There were only two Hungarian pipelines that try to serve industrial needs. One of them, \magyarlanc{} \cite{magyaralanc}, was designed for industrial applications offering several desirable features such as software quality, speed, memory efficiency, and customizability. However, despite being used in commercial applications in the real world, it has not been maintained for several years and lacks integration with the Python ecosystem. The other pipeline, called \emtsv{} \cite{emtsv1,emtsv2,emtsv4}, aimed to integrate existing NLP toolkits into a single application, but neither computational efficiency nor developer ergonomics were the main goals of the project. Additionally, while \magyarlanc{} natively uses the universal morphosyntactic features, \emtsv{} can only do this through conversion. Both pipelines use dependency annotation that is incompatible with \universaldependencies, furthermore, none of them can utilize word embeddings or large language models, which have become increasingly important in recent years.

In contrast, the development of \huspacy{} placed emphasis not only on accuracy, but also on software ergonomics, while also adhering to the international standards established by Nivre et al. \cite{nivre-etal-2020-universal}. Moreover, it is built on \spacy, enabling users to access its full functionality with ease. One significant drawback of this tool is the lack of precise annotations for lemmata, entities and dependencies syntax.

To fulfill the industrial requirements of text processing pipelines, this work is built on the \universaldependencies{} annotation schema and our models are implemented in \spacy{} by extending \huspacy’s text processing model. The detailed documentation, intuitive API, high speed and accuracy of these tools make them an optimal choice for building high-performing NLP models. Additionally, \huspacy{} utilizes non UD compatible corpora as well, which allows for a comprehensive analysis of Hungarian texts.

\section{Methods} \label{sec:methods}
\subsection{\huspacy\textquotesingle s Internals} \label{sec:methods:internals}

\huspacy’s main strength lies in the clever usage of available Hungarian linguistic resources and its multi-task learning capabilities inherited from \spacy. Its machine learning approach can be summarized as ``embed, encode, attend, predict''
%\footnote{Details are provided here: \\ \url{https://explosion.ai/blog/deep-learning-formula-nlp}} 
shown in Figure \ref{fig:architecture} and detailed by \cite{spacy-parser,huspacy:2021}. Tokens are first embedded through the combination of lexical attributes and word vectors, then context encoding is performed by stacked CNN \cite{cnn} layers\footnote{These steps are usually referred to as the Tok2Vec layers.}. Finally, task specific layers are used parallelly in a multi-task learning setup.

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{architecture_1.pdf}
\caption{The ``embed, encode, attend, predict'' architecture of \spacy}
\label{fig:architecture}
\end{figure}

Orosz et al. \cite{huspacy:2021} used a three step approach for fully utilizing annotated Hungarian datasets. First, they pre-train the tagger, the lemmatizer and the sentence boundary detection components on a silver standard \ud{} annotated corpus (cf. \cite{vincze-etal-2017-universal}). Then, the Tok2Vec layers of this model are reused by both the NER and the parsing components: the dependency parser and the morphosyntactic taggers are fine-tuned on the \udhu{} dataset, the lemmatizer is trained on the entire \szegedcorpus, while the entity recognizer is further trained on the combination of the \nerkor{} and the \szegedner{} datasets.

\subsection{Improving on the Underlying Language Models} \label{sec:methods:underlying}

\huspacy\textquotesingle s model is built on \wordtovec{} \cite{word2vec} word embeddings, which are known to have limitations in providing meaningful representations for out-of-vocabulary words. This is particularly problematic for morphology-related tasks in agglutinative languages. To enhance this simple approach, a more fine-grained method that uses sub-word embeddings can be employed. \fasttext{} \cite{bojanowski2017enriching} is a widely-used extension of \wordtovec{} that learns sub-token embeddings. In this study, we utilized \floret\footnote{\url{https://explosion.ai/blog/floret-vectors}} which is a \spacy-compatible fork of \fasttext. To train new word vectors, we used the Hungarian Webcorpus 2.0 \cite{hubert}. Two sets of word embeddings were constructed: a 100-dimensional and a 300-dimensional one.

In recent years, there has been a growing interest in transformer-based large language models (LLM), as evidenced by their high performance in text processing models (e.g. \cite{hubert,enevoldsen2021dacy}). With the advent of \spacy\textquotesingle s native support for such architectures and the availability of pre-trained language models for Hungarian, it is now possible to train transformer-based NLP pipelines for Hungarian. Our research is based on two widely used LLMs that provide support for Hungarian. One of these is \hubert{} \cite{hubert}, which has a \bert-like architecture and was trained using monolingual data. The other model is \xlmroberta{}, which has a much larger capacity compared to the former model and was trained on multilingual corpora.

\subsection{Pipeline Component Enhancements} \label{sec:methods:enchancements}

In addition to the use of more powerful language models, we propose fundamental changes to the lemmatization and dependency parsing models, as well as minor improvements to the entity recognizer.

\huspacy\textquotesingle s lemmatizer has been replaced by a new edit-tree-based architecture, recently available in the \spacy{} framework\footnote{\url{https://explosion.ai/blog/edit-tree-lemmatizer}}. This new model builds on the foundations laid out by Müller et al. \cite{muller-etal-2015-joint} (called the \lemming{} model), but has minor differences from it. On the one hand, this reimplementation fully utilizes the framework’s multi-task learning capabilities, which means that the lemmatizer is not only co-trained with PoS and morphological tagging, but also with sentence boundary detection. On the other hand, \spacy\textquotesingle s version lacks standard support for morphological lexicons which \lemming{} benefited from.

We have improved this model in two steps. 
\begin{enumerate*}
    \item A simple dictionary learning method is put in place to memorize frequent \textit{(token, tag, lemma)} triplets of the training data which are then used at prediction time to retrieve the roots of words. 
    \item A common weakness of Hungarian lemmatization methods is addressed. Computing the lemmata of sentence-starting tokens can be challenging for non-proper nouns, as their roots are always lowercase. Thus, we force the model to use the true casing of such words. For example, when computing the root of the sentence starting \textit{Ezzel} `with this' token, our method checks its PoS tag (that is ideally PRON) first, so that it can use the lowercase wordform for generating and looking up edit-trees.
\end{enumerate*}

Moving on, the dependency syntax annotation component is replaced with a model that has higher accuracy for many languages. Although \spacy\textquotesingle s built-in transition-based parser \cite{spacy-parser} has high throughput, it falls short on providing accurate predictions. Graph-based architectures are known to have good performance for dependency parsing (e.g. \cite{altintacs2023improving}), making such methods good enhancement candidates. Furthermore, a \spacy-compatible implementation of Dozat and Manning’s model \cite{biaffine} (referred to as the \biaffine{} parser) has recently been made available, thus we could easily utilize it in our experiments.

Finally, the named entity recognizer has been fine-tuned to provide more accurate entity annotations. This was primarily achieved by using beam-search in addition to the transition-based NER module.

\section{Experiments and Results} \label{sec:experiments}

This section presents the results of several experiments that demonstrate the improvements of our changes and show competitive results compared to well-established baselines. We evaluated pipelines developed on datasets used by the creators of \huspacy: the Hungarian part of the \universaldependencies{} corpus\footnote{Experiments are performed at the v2.10 revision.} was utilized to benchmark the sentence boundary detector, the lemmatizer, the PoS and morphological taggers, and the dependency parser, while the entity recognizer is benchmarked on the combination of the \nerkor{} and the \szegedner{} corpora (similar to \cite{huspacy:2021} and \cite{nerkoreval}). To account for the instability of \spacy’s training process we report the maximum result of three independent runs.

\subsection{Evaluation of Architecture Improvements} \label{sec:experiments:eval}

The lemmatization accuracy of the original model has been greatly improved through a number of steps discussed in Section \ref{sec:methods:enchancements}. As evidenced in Table \ref{table:lemma}, incorporation of the new neural architecture along with sub-word embeddings produced significant improvements. Furthermore, changing the default behavior of the edit-tree lemmatizer by allowing it to evaluate more than one candidate (see the row \texttt{topk=3}) also resulted in a slightly better performance. In addition, the integration of true-casing led to a considerable improvement, and the use of lemma dictionaries also significantly improved lemmatization scores.

\newcolumntype{R}{>{\centering\arraybackslash}X}
\begin{table}[H] % table 1
	\begin{center}
		\caption{Lemmatization accuracy on the \udhu{} test set of different ablation settings. Rows marked with a ``+'' indicate a new feature added on top of the previous ones. \texttt{topk} is a hyperparameter of the lemmatization model controlling the number of edit-trees considered to be evaluated.}
		\begin{tabularx}{5.75cm}{l R}
			\toprule
			                       & \multicolumn{1}{c}{Lemma Accuracy} \\
			\midrule
			\huspacy               & 95.53\%                            \\
			+ Edit-tree lemmatizer & 95.90\%                            \\
			+ \floret{} 300d vectors    & 96.76\%                            \\
			+ \texttt{topk=3}      & 97.01\%                            \\
			+ True-casing          & 97.30\%                            \\
			+ Learned dictionary   & 97.58\%                            \\
			\bottomrule
		\end{tabularx}
		\vspace{1em}
		\label{table:lemma}
	\end{center}
	\vspace{-2em}
\end{table}

Entity recognition tasks often encounter a challenge in the form of a considerable number of out-of-vocabulary tokens, leading to decreased performance. However, the utilization of \floret{} vectors has proven to be effective in addressing this issue, as indicated by the results in Table \ref{table:ner}. Additionally, the use of beam search allowed the model to take prediction history into account, which slightly improved its efficiency.

\newcolumntype{R}{>{\centering\arraybackslash}X}
\begin{table} [H]% table 2
	\begin{center}
		\caption{Evaluation of the entity recognition model improvements on the combination of the \szegedner{} and \nerkor{} corpora.  The rows starting with ``+'' signify the inclusion of a new feature in addition to the existing ones.}
		\begin{tabularx}{5cm}{l R}
			\toprule
			                    & \multicolumn{1}{c}{NER \fone-score} \\
			\midrule
			\huspacy            & 83.68                               \\
			+ \floret{} 300d vectors & 85.53                               \\
			+ Beam search       & 85.99                               \\
			\bottomrule
		\end{tabularx}
		\vspace{1em}
		\label{table:ner}
	\end{center}
	\vspace{-2em}
\end{table}

The results in Table \ref{table:parser} indicate that the improved text representations and the new parsing architecture offer substantial improvements over \huspacy\textquotesingle s outcomes. However, it is worth noting that \spacy’s CNN-based base model is not fully compatible with the \biaffine{} parser’s architecture. Therefore, parsing improvements were benchmarked on top of a transformer-based encoder architecture using \hubert. The results show that the use of \floret{} vectors is beneficial to predict morphosyntactic characteristics and dependency relations, while the use of \hubert-based text representations substantially improves performance across all subtasks. Furthermore, the \biaffine{} parser significantly outperforms its transition-based counterpart, as evidenced by its better attachment scores.

\newcolumntype{R}{>{\centering\arraybackslash}X}
\begin{table}[H] % table 3
	\begin{center}
		\caption{Evaluation of text parsing improvements on the \udhu{} test set. ``+'' indicate a new feature added on top of the existing ones.}
		\begin{tabularx}{9cm}{l R R R R}
			\toprule
			                    & \multicolumn{1}{c}{\makecell{PoS \\ Acc.}} & \multicolumn{1}{c}{\makecell{Morph. \\ Acc.}} & \multicolumn{1}{c}{UAS}   & \multicolumn{1}{c}{LAS}   \\
			\midrule
			\huspacy             & 96.58\%  & 93.23\%     & 79.39 & 74.22 \\
			\huspacy{} + \floret{} 300d vectors & 96.55\%  & 93.93\%     & 80.36 & 74.89 \\
			\huspacy{} + \hubert           & 98.10\%  & 96.97\%     & 89.95 & 83.94 \\
			+ \biaffine{} parser  & 98.10\%  & 96.97\%     & 90.31 & 87.23 \\
			\bottomrule
		\end{tabularx}
		\vspace{1em}
		\label{table:parser}
	\end{center}
	\vspace{-2em}
\end{table}

\subsection{Comparison with the State-of-the-Art} \label{sec:experiments:sota} % "If the first word can stand alone, the second word should be capitalized." + az of és a the stopszavak, szóval az kisbetű, a wikipédia meg az egészet külön írja

In addition to parsing and tagging correctness, resource consumption is an important consideration for industrial NLP applications. Therefore, following the approach of Orosz et al. \cite{huspacy:2021} we conducted a benchmark study to compare both the accuracy and memory usage as well as the throughput of our models with text processing tools available for Hungarian.

\newcolumntype{R}{>{\centering\arraybackslash}X}
\begin{table}[H]% table 4
	\begin{center}
		\caption{Text parsing accuracy of the novel pipelines compared to \huspacy, \stanza, \udify, \trankit{} and \emtsv. Results for non-comparable models are shown in italics.}
		\begin{tabularx}{\textwidth}{l R R R R R R R}
			\toprule
			                  & \multicolumn{1}{c}{\makecell{Sent. \\ \fone-score}}    & \multicolumn{1}{c}{\makecell{PoS \\ Acc.}}         & \multicolumn{1}{c}{\makecell{Morph. \\ Acc.}}      & \multicolumn{1}{c}{\makecell{Lemma \\ Acc.}}       & \multicolumn{1}{c}{UAS}                     & \multicolumn{1}{c}{LAS}                     & \multicolumn{1}{c}{\makecell{NER \\ \fone-score}}      \\
			\midrule
			\textit{\emtsv}   & \textit{98.11} & \textit{89.19\%} & \textit{87.95\%} & \textit{96.16\%} & \textit{--}              & \textit{--}              & \textbf{92.99} \\
			\textit{\trankit} & \textit{98.00} & \textit{97.49\%} & \textit{95.23\%} & \textit{94.45\%} & \textit{\textbf{91.31}} & \textit{\textbf{87.78}} & \textit{--}              \\
			\udify            & --              & 96.15\%          & 90.54\%          & 88.70\%          & 88.03                   & 83.92                   & --              \\
			\stanza           & 97.77          & 96.12\%          & 93.58\%          & 94.68\%          & 84.05                   & 78.75                   & 83.75          \\
			\midrule
			\huspacy          & 97.54          & 96.58\%          & 93.23\%          & 95.53\%          & 79.39                   & 74.22                   & 83.68          \\
			\huspacymd        & 97.88          & 96.26\%          & 93.29\%          & 97.38\%          & 79.25                   & 73.99                   & 85.35          \\
			\huspacylg        & 98.33          & 96.91\%          & 93.93\%          & 97.58\%          & 79.75                   & 74.78                   & 85.99          \\
			\huspacytrf       & 99.33          & \textbf{98.10\%} & \textbf{96.97\%} & 98.79\%          & 90.31                   & 87.23                   & 91.35          \\
			\huspacytrfxl     & \textbf{99.67} & 97.79\%          & 96.53\%          & \textbf{98.90\%} & 90.22                   & 86.67                   & 91.84          \\
			\bottomrule
		\end{tabularx}
		% \vspace{1em}
		\label{table:compare}
	\end{center}
	\vspace{-3em}
\end{table}

\newcolumntype{R}{>{\raggedleft\arraybackslash}X}
\begin{table}[H] % table 5
	\begin{center}
		\caption[Resource usage of the new models and state-of-the-art of text processing tools available for Hungarian.]{Resource usage\protect\footnotemark{} of the new models and state-of-the-art of text processing tools available for Hungarian. Throughput is measured as the average number of processed tokens per second, while memory usage columns records the peak value of each tool.}
		% https://tex.stackexchange.com/questions/10181/using-footnote-in-a-figures-caption
		\begin{tabularx}{8.5cm}{l R R R}
  \toprule
        &                         & \multicolumn{1}{l}{Throughput}  & \multicolumn{1}{c}{\multirow{2}{*}{\makecell{Memory Usage\\(GB)}}} \\
        & \multicolumn{1}{r}{CPU} & \multicolumn{1}{r}{GPU}         & \multicolumn{1}{c}{}                                               \\
  \midrule
			\emtsv        & 113            & --             & 3.9               \\
			\trankit      & 434            & 2119           & 3.7               \\
			\udify        & 129            & 475            & 3.2               \\
			\stanza       & 30             & 395            & 5.3               \\
			\midrule
			\huspacy      & 1525           & 6697           & 3.5               \\
			\huspacymd    & 2652           & 3195           & 1.4               \\
			\huspacylg    & 847            & 3128           & 3.2               \\
			\huspacytrf   & 273            & 2605           & 4.8               \\
			\huspacytrfxl & 82             & 2353           & 18.9              \\
			\bottomrule
		\end{tabularx}
		% \vspace{1em}
		\label{table:speed}
	\end{center}
	% \vspace{-3em}
\end{table}
\footnotetext{All benchmarks are run on the same environment having AMD EPYC 7F72 CPUs and NVIDIA A100 GPUs.}

First of all, an important result of this study is a base model (referred to as \huspacylg), which achieves a good balance between accuracy and resource usage as seen in Tables \ref{table:compare} and \ref{table:speed}. This pipeline is built on top of the 300d \floret{} vectors and incorporates all the enhancements described above, except for the new parser. Evaluation data demonstrates that the \huspacylg{} pipeline consistently outperforms \stanza{} in all tasks except syntactic dependency relation prediction, which can be explained by the superior parsing model of the latter tool.

We present the results of a medium-sized model (\huspacymd) as well that is a reduced version of the \huspacylg{} pipeline utilizing the smaller (100d) word embeddings. Surprisingly, the \huspacymd{} pipeline delivers performance similar to that of the larger model. Furthermore, the medium-sized model achieves scores comparable to or higher than those of \huspacy, despite requiring half the memory and exhibiting much higher throughput on CPU.

Transformer-based pipelines using the graph based dependency parser have the highest scores across all language analysis tasks. Remarkably, despite its smaller capacity, the model based on \hubert{} (\huspacytrf) achieves the highest attachment scores for dependency parsing, while the one using \xlmroberta{} (\huspacytrfxl) provides slightly more accurate PoS tags and named entities.

It is important to consider that not all third-party pipelines in Table \ref{table:compare} are directly comparable to our results, due to differences in the versions of the \udhu{} dataset used to train and evaluate their models. To ensure a fair comparison, \stanza{} and \udify{} have been retrained. On the other hand, we obtained the results of \trankit{} from \cite{van2021trankit} since it would be a demanding task to fine-tune this model. Furthermore, the results of \emtsv\textquotesingle s text parsing components \cite{orosz-novak-2013-purepos,novak2014new,novak-etal-2016-new} cannot be deemed reliable either (cf. \cite{huspacy:2021}), since its components use a different train-test split of the \szegedcorpus. However, this tool’s entity recognition module (\embert{} \cite{embert}) was evaluated by Simon et al. \cite{nerkoreval} using the same settings as in our paper, thus we rely on their assessment. Additionally, state-of-the-art results are also shown in Table \ref{table:compare}. With regard to highest dependency parsing scores, the results of the multilingual \trankit{} system are produced by a parsing model similar to that of ours. As for named entity recognition, \embert{} attains the best \fone{} scores by utilizing a Viterbi encoder that eliminates invalid label sequences from the outputs of the underlying model.

Regarding computational requirements, Table \ref{table:speed} presents findings that demonstrate how \floret{} embeddings can effectively decrease the memory usage of models without compromising their accuracy and throughput. However, it is apparent that enhancing pipeline accuracy frequently results in slower processing speed, as can be observed from the \huspacylg, \huspacytrf{} and \huspacytrfxl{} models. Additionally, our tests also showed that most of the readily available NLP pipelines are not adequately optimized to handle large workloads, which is evident from their low throughput values.

\section{Conclusion} \label{sec:conclusion}

This paper has introduced new industrial-grade text processing pipelines for Hungarian and presented a thorough evaluation showing their (close to) state-of-the-art performance. We have shown that new architectures for lemmatization and dependency parsing and the use of improved text representation models significantly improve the accuracy of \huspacy. The presented models have not only demonstrated high performance in all text preprocessing steps, but the resource consumption of three of our models’ (\huspacymd, \huspacylg, \huspacytrf) makes them suitable for solving practical problems. All of our experiments are reproducible and the models are freely available under a permissive license.%\footnote{\url{https://github.com/huspacy/huspacy}}. %NOTE: első kör miatt hide

For future work, we consider the following areas of improvement.
\begin{enumerate*}
    \item Transformer\-based pipelines are optimized for accuracy but this could limit their usability due to reduced computational efficiency. We would like to investigate optimizing their size to enhance their resource usage. 
    \item We would like to include more silver standard data to further improve the parsing and tagging scores, as the corpus used to train and evaluate text parsing components is limited in size. 
    \item Our models are mostly trained on news\hyp{}related corpora, which makes user-generated text processing a difficult task. In order to address this challenge, we intend to integrate automatic data augmentation into the training process as a solution.
\end{enumerate*}

\subsubsection*{Acknowledgments.} The authors would like to thank Gábor Berend for his valuable suggestions.
\huspacy{} research and development is supported by the European Union project RRF-2.3.1-21-2022-00004 within the framework of the Artificial Intelligence National Laboratory. % source: idei mszny

% Bibliography
% \todo{Gyuri: arxivos hivatkozast csak akkor hasznaljunk, ha nincs igazi. Szoval meg kellene keresni minden ilyen cikk eredetijet es arra hivatkozni}
% \todo{Gyuri: hivatkozasok cimeinek nagybetusitese ahol kell, ui a bibtex mindent automatikusan kisbetusit}
\bibliographystyle{splncs04}
\bibliography{paper}

\end{document}
